### 2022-04-12 extras/bookdown/index.Rmd
initialize

--- 
author: Alexandros Rekkas
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
biblio-style: apalike
bibliography: [ "references.bib" ]
link-citations: true
csl: jce.csl
---

% Pandoc header
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\usepackage{amsmath}
\doublespacing
\usepackage{amssymb}
\usepackage{caption}
\usepackage{booktabs}
\date{}
\newcommand\given[1][]{\:#1\vert\:}
\newcommand{\indep}{\perp \!\!\! \perp}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(SmoothHte)
library(rms)
library(here)

d <- function(x, decimals = 4) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}

inline_hook <- function(x) {
  if (is.numeric(x)) {
    if (abs(x - round(x)) < .Machine$double.eps) {
      formatted <- format(x, digits = 1, big.mark = ",")
    } else {
      formatted <- format(x, digits = 1 ,nsmall = 3, big.mark = ",")
    }
  } else {
    formatted <- x
  }

  return(formatted)
}

knit_hooks$set(inline = inline_hook)

set.seed(19910930)
```

\doublespacing 
\linenumbers

# Introduction
Predictive approaches to heterogeneity of treatment effects (HTE) aim at the
development of models predicting either individualized effects or which of two
(or more) treatments is better for an individual [@Varadhan2013]. In prior work,
we divided such methods in three broader categories based on the reference class
used for defining patient similarity when making individualized predictions or
recommendations [@Rekkas2020]. Risk-modeling approaches use prediction of
baseline risk as the reference; treatment effect modeling approaches also model
treatment-covariate interactions, in addition to risk factors; optimal treatment
regime approaches focus on developing treatment assignment rules and rely
heavily on modeling treatment effect modifiers.

Risk-modeling approaches to predictive HTE analyses provide a viable option in
the absence of well-established treatment effect modifier [@Kent2019;
@PathEnE]. In simulations, modeling treatment-covariate interactions, often led
to miscalibrated predictions of absolute benefit, contrary to risk-based
methods, despite their weaker discrimination of benefit in the presence of true
effect modifiers [@vanKlaveren2019]. Most often, risk-modeling approaches are
carried out in two steps: first a risk prediction model is developed externally
or internally on the entire RCT population, “blinded” to treatment; then the RCT
population is stratified using this prediction model to evaluate risk-based
treatment effect variation [@Kent2010]. This approach identified substantial
absolute treatment effect differences between low-risk and high-risk patients in
a re-analysis of 32 large trials [@Kent2016]. However, even though estimates at
the risk subgroup level may be accurate, these estimates may not apply to
individual patients.

In the current simulation study, we aim to summarize and compare different
risk-based models for predicting treatment effects. We simulate different
relations between baseline risk and treatment effects and also consider
potential harms of treatment. We illustrate the different models by a case study
of predicting individualized effects of treatment for acute myocardial
infarction (MI) in a large RCT.

# Methods

## Notation

We observe RCT data $(Z, X, Y)$, where for each patient $Z_i= 0, 1$ is the
treatment status, $Y_i = 0, 1$ is the observed outcome and $X_i$ is a set of
measured covariates. Let $\{Y_i(z), z=0, 1\}$ denote the unobservable potential
outcomes. We observe $Y_i = Z_iY_i(1) + (1 - Z_i)Y_i(0)$. We are interested in
predicting the conditional average treatment effect (CATE), 
$$\tau(x) = E\{Y(0) - Y(1)|X=x\}$$ 
Assuming that $(Z, X, Y)$ is a random sample from the target population and that
$\big(Y(0), Y(1)\big)\indep Z|X$, as we are in the RCT setting, we can predict
CATE from
\begin{align*}
\tau(x) &= E\{Y(0)\given X=x\}-E\{Y(1)\given X=x\}\\
&=E\{Y\given X=x, Z=0\}-E\{Y\given X=x, Z=1\}
\end{align*}

## Simulation scenarios

We simulated a typical RCT, comparing equally-sized treatment and control arms
in terms of a binary outcome. For each patient we generated 8 baseline
covariates $x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim
B(1,0.2)$. Outcomes in the control arm were generated from Bernoulli variables
with true probabilities following a logistic regression model including all
baseline covariates, i.e.
$P(Y(0)=1\given X=x) = \text{expit}(lp_0) = e^{lp_0}/(1+e^{lp_0})$,
with $lp_0=lp_0(x)=x^t\beta$.
In the base scenarios coefficient values $\beta$ were
such, that the AUC of the logistic regression model was 0.75 and the event rate
in the control arm was $20\%$.

Outcomes in the treatment arm were first generated using 3 simple scenarios:
absent (OR = 1), moderate (OR = 0.8) or strong (OR = 0.5) constant relative
treatment effect. We then introduced linear, quadratic and non-monotonic
deviations from constant treatment effects using:
$$lp_1 = \gamma_2(lp_0-c)^2 + \gamma_1(lp_0-c) + \gamma_0, $$
where $lp_1$ is the true linear predictor in the treatment arm, so that
$P(Y(1)=1\given X=x) = \text{expit}(lp_1)$.
Finally, we incorporated constant absolute harms for all treated patients, such
that
$P(Y(1)=1|X=x) = \text{expit}(lp_1) + \text{harm}$.

The sample size for the base scenarios was set to 4,250 (80\% power for the
detection of a marginal OR of 0.8 with the standard alpha of 5\%). We evaluated
the effect of smaller or larger sample sizes of 1,063 and 17,000,
respectively. We also evaluated the effect of risk model discriminative ability,
adjusting the baseline covariate coefficients, such that the AUC of the
regression model in the control arm was 0.65 and 0.85, respectively.

These settings resulted in a simulation study of 648 scenarios covering the
observed HTE in 32 large trials as well as many other potential variations of
risk-based treatment effect (Supplement, Sections 2 and 3) [@Kent2016].

## Individualized risk-based benefit predictions

In each simulation run we internally developed a prediction model on the entire
population, using a logistic regression with main effects for all baseline
covariates and treatment assignment. Individual risk predictions were derived by
setting treatment assignment to 0. Another approach would be to derive the
prediction model solely on the control patients; however, it has been shown to
lead to biased benefit predictions [@vanKlaveren2019; @Burke2014; @Abadie2018].

A *stratified HTE method* has been suggested as an alternative to traditional
subgroup analyses [@Kent2019; @PathEnE]. Patients are stratified into
equally-sized risk strata—in this case based on risk quartiles. Absolute
treatment effects within risk strata are estimated by the difference in event
rate between control and treatment arm patients. We considered this approach as
a reference, expecting it to perform worse than the other candidates, as its
objective is to provide an illustration of HTE rather than to optimize
individualized benefit predictions.

Second, we considered a model which assumes *constant relative treatment effect*
(constant odds ratio). Hence, absolute benefit is predicted from
$\tau(x;\hat{\beta}) = \text{expit}(\hat{lp}_0) - \text{expit}(\hat{lp}_0+\delta_0)$, where
$\delta_0$ is the log of the assumed constant odds ratio and
$\hat{lp}_0 = \hat{lp}_0(x;\hat{\beta}) = x^t\hat{\beta}$
the linear predictor of the estimated baseline risk model.

Third, we considered a logistic regression model including treatment, the
prognostic index, and their linear interaction. Absolute benefit is then
estimated from
$\tau(x;\hat{\beta}) = \text{expit}(\delta_0+\delta_1\hat{lp}_0) - \text{expit}(\delta_0+\delta_2+(\delta_1+\delta_3)\hat{lp}_0)$
We will refer to this method as the *linear interaction* approach.

Fourth, we used *restricted cubic splines* (RCS) to relax the linearity
assumption on the effect of the linear predictor [@Harrell1988]. We considered
splines with 3 (RCS-3), 4 (RCS-4) and 5 (RCS-5) knots to compare models with
different levels of flexibility.

Finally, we considered an adaptive approach using Akaike’s Information Criterion
(AIC) for model selection. More specifically, we ranked the constant relative
treatment effect model, the linear interaction model, and the RCS models with 3,
4, and 5 knots based on their AIC and selected the one with the lowest
value. The extra degrees of freedom were 1 (linear interaction), 2, 3 and 4 (RCS
models) for these increasingly complex interactions with the treatment effect.

## Evaluation metrics
We evaluated the predictive accuracy of the considered methods by the root mean
squared error (RMSE):

$$\text{RMSE}=\sqrt{\frac{1}{n}\sum_{i=1}^n\big(\tau(x_i) - \hat{\tau}(x_i)\big)^2}$$

We compared the discriminative ability of the methods under study using
c-for-benefit and the integrated calibration index (ICI) for
benefit (Supplement, Section 6) [@vanKlaveren2018].

For each scenario we performed 500 replications, within which all the considered
models were fitted. We simulated a super-population of size 500,000 for each
scenario within which we calculated RMSE and discrimination and calibration for
benefit of all the models in each replication.

## Empirical illustration

```{r, echo=FALSE, warning=FALSE, message=FALSE}
load(here::here("data/raw/gusto.rda"))
gusto <- gusto %>%
  tibble() %>%
  filter(!is.na(tpa))

gusto <- gusto %>%
  tibble() %>%
  filter(tx != "SK+tPA") %>%
  rename(
    "outcome" = "day30",
    "treatment" = "tpa"
  )

treatmentArms <- gusto %>%
  group_by(treatment) %>%
  summarise(n = n())
```

We demonstrated the different methods using `r nrow(gusto)` patients with acute
myocardial infarction (MI) included in the GUSTO-I trial.  `r treatmentArms %>%
filter(treatment == 1) %>% pull(n)` patients were randomized to tissue
plasminogen activator (tPA) treatment and `r treatmentArms %>% filter(treatment
== 0) %>% pull(n)` were randomized to streptokinase. The outcome of interest
was 30-day mortality (total of `r sum(gusto$outcome)` events), recorded for all
patients. In line with previous analyses [@Califf1997; @Steyerberg2000], we
fitted a logistic regression model with 6 baseline covariates, i.e. age, Killip
class, systolic blood pressure, heart rate, an indicator of previous MI, and the
location of MI, to predict 30-day mortality risk (Supplement, Section 8).

# Results

## Simulations

```{r adaptive, echo=FALSE, warning=FALSE, message=FALSE}
adaptiveSelections <- readr::read_csv(here::here("data/processed/adaptiveSelections.csv"))
rmseDistribution <- readr::read_csv(here::here("data/processed/rmseDistribution.csv"))
aucDistribution <- readr::read_csv(here::here("data/processed/discriminationDistribution.csv"))
```

```{r rmse, echo=FALSE, warning=FALSE, message=FALSE}
analysisIds <- readr::read_csv(here::here("data/processed/analysisIds.csv"))
rmse <- readr::read_csv(here::here("data/processed/rmse.csv"))
selectedScenario <- analysisIds %>%
  dplyr::filter(
    base == "moderate",
    type == "constant",
    sampleSize == 4250,
    auc == 0.75,
    harm == "absent"
  ) %>%
  dplyr::pull(scenario)

rmseBaseCase <- rmse %>% dplyr::filter(scenarioId == selectedScenario)

selectedScenario <- analysisIds %>%
  dplyr::filter(
    base == "moderate",
    type == "linear-high",
    sampleSize == 4250,
    auc == 0.75,
    harm == "absent"
  ) %>%
  dplyr::pull(scenario)

rmseLinearInteraction <- rmse %>% dplyr::filter(scenarioId == selectedScenario)

selectedScenario <- analysisIds %>%
  dplyr::filter(
    base == "moderate",
    type == "quadratic-high",
    sampleSize == 4250,
    auc == 0.75,
    harm == "absent"
  ) %>%
  dplyr::pull(scenario)

rmseQuadratic <- rmse %>% dplyr::filter(scenarioId == selectedScenario)

selectedScenario <- analysisIds %>%
  dplyr::filter(
    base == "moderate",
    type == "non-monotonic",
    sampleSize == 4250,
    auc == 0.75,
    harm == "absent"
  ) %>%
  dplyr::pull(scenario)

rmseNonMonotonic <- rmse %>% dplyr::filter(scenarioId == selectedScenario)

selectedScenario <- analysisIds %>%
  dplyr::filter(
    base == "moderate",
    type == "non-monotonic",
    sampleSize == 4250,
    auc == 0.75,
    harm == "strong-positive"
  ) %>%
  dplyr::pull(scenario)

rmseNonMonotonicHarm <- rmse %>% dplyr::filter(scenarioId == selectedScenario)
```

The constant treatment effect approach outperformed other approaches in the base
case scenario (N = `r 4250`; OR = 0.8; AUC= 0.75; no absolute treatment harm)
with a true constant treatment effect (median RMSE: constant treatment effect 
`r median(rmseBaseCase$constant_treatment_effect)`;
linear interaction `r median(rmseBaseCase$linear_predictor)`;
RCS-3 `r median(rmseBaseCase$rcs_3_knots)`). 
The linear interaction model was optimal under true linear deviations 
(median RMSE: constant treatment effect 
`r median(rmseLinearInteraction$constant_treatment_effect)`;
linear interaction `r median(rmseLinearInteraction$linear_predictor)`;
RCS-3 `r median(rmseLinearInteraction$rcs_3_knots)`; Figure \ref{fig:rmsebase} panels A-C)
and even in the presence of true quadratic deviations
(median RMSE: constant treatment effect 
`r median(rmseQuadratic$constant_treatment_effect)`;
linear interaction `r median(rmseQuadratic$linear_predictor)`;
RCS-3 `r median(rmseQuadratic$rcs_3_knots)`; Figure \ref{fig:rmsebase} panels A-C)
from a constant relative treatment effect. With non-monotonic deviations, RCS-3
slightly outperformed the linear interaction model
(Median RMSE: linear interaction `r median(rmseNonMonotonic$linear_predictor)`;
RCS-3 `r median(rmseNonMonotonic$rcs_3_knots)`; Figure \ref{fig:rmsebase} panel D).
With strong treatment-related harms the results were very similar in most
scenarios (Figure \ref{fig:rmsebase} panels A-C). Under non-monotonic deviations
the optimal performance of RCS-3 was more pronounced
(Median RMSE: linear interaction `r median(rmseNonMonotonicHarm$linear_predictor)`;
RCS-3 `r median(rmseNonMonotonicHarm$rcs_3_knots)`; Figure \ref{fig:rmsebase} panel D).
A stronger average treatment effect (OR=0.5) led to larger absolute benefit
predictions and consequently to larger RMSE for all approaches, but the relative
differences between different approaches were similar to the base case scenario
(Supplement, Figure S10).

```{r rmsebase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated from a simulated super-population of size 500,000. The scenario with true constant relative treatment effect (panel A) had a true prediction AUC of 0.75 and sample size of 4250. The RMSE is also presented for strong linear (panel B), strong quadratic (panel C), and non-monotonic (panel D) from constant relative treatment effects. Panels on the right side present the true relations between baseline risk (x-axis) and absolute treatment benefit (y-axis). The 2.5, 25, 50, 75, and 97.5 percentiles of the risk distribution are expressed by the boxplot on the top. The 2.5, 25, 50, 75, and 97.5 percentiles of the true benefit distributions are expressed by the boxplots on the side of the right-handside panel.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_base.tiff")))
# knitr::include_graphics(here::here("figures/rmse_base.png"))
```

The adaptive approach had limited loss of performance in terms of the median
RMSE to the best-performing method in each scenario. However, compared to the
best-performing approach, its RMSE was more variable in scenarios with linear
and non-monotonic deviations, especially when also including moderate or strong
treatment-related harms. On closer inspection, we found that this behavior was
caused by selecting the constant treatment effect model in a substantial
proportion of the replications (Supplement, Figure S3).

```{r rmseSamplesize, echo=FALSE, warning=FALSE, message=FALSE}
selectedScenario <- analysisIds %>%
  dplyr::filter(
    base == "moderate",
    type == "constant",
    sampleSize == 17000,
    auc == 0.75,
    harm == "absent"
  ) %>%
  dplyr::pull(scenario)

rmseBaseCaseSampleSize <- rmse %>% dplyr::filter(scenarioId == selectedScenario)

selectedScenario <- analysisIds %>%
  dplyr::filter(
    base == "moderate",
    type == "linear-high",
    sampleSize == 17000,
    auc == 0.75,
    harm == "absent"
  ) %>%
  dplyr::pull(scenario)

rmseLinearInteractionSampleSize <- rmse %>% dplyr::filter(scenarioId == selectedScenario)

selectedScenario <- analysisIds %>%
  dplyr::filter(
    base == "moderate",
    type == "quadratic-high",
    sampleSize == 17000,
    auc == 0.75,
    harm == "absent"
  ) %>%
  dplyr::pull(scenario)

rmseQuadraticSampleSize <- rmse %>% dplyr::filter(scenarioId == selectedScenario)

selectedScenario <- analysisIds %>%
  dplyr::filter(
    base == "moderate",
    type == "non-monotonic",
    sampleSize == 17000,
    auc == 0.75,
    harm == "absent"
  ) %>%
  dplyr::pull(scenario)

rmseNonMonotonicSampleSize <- rmse %>% dplyr::filter(scenarioId == selectedScenario)
```

Increasing the sample size to `r 17000` favored RCS-3 the most. The difference in
performance with the linear interaction approach was more limited in settings
with a constant treatment effect 
(Median RMSE: linear interaction `r median(rmseBaseCaseSampleSize$linear_predictor)`;
RCS-3 `r median(rmseBaseCaseSampleSize$rcs_3_knots)`) and with a true linear interaction
(Median RMSE: linear interaction `r median(rmseLinearInteractionSampleSize$linear_predictor)`;
RCS-3 `r median(rmseLinearInteractionSampleSize$rcs_3_knots)`).
and more emphasized in settings with strong quadratic deviations 
(Median RMSE: linear interaction `r median(rmseQuadraticSampleSize$linear_predictor)`;
RCS-3 `r median(rmseQuadraticSampleSize$rcs_3_knots)`) and non-monotonic deviations
(Median RMSE: linear interaction `r median(rmseNonMonotonicSampleSize$linear_predictor)`;
RCS-3 `r median(rmseNonMonotonicSampleSize$rcs_3_knots)`).
Due to the large sample size, the RMSE of the adaptive approach was even more
similar to the best-performing method, and the constant relative treatment
effect model was less often wrongly selected (Supplement, Figure S4).

```{r rmsesamplesize, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in simulated samples of size 17,000 rather than 4,250 in Figure \\ref{fig:rmsebase}. RMSE was calculated on a super-population of size 500,000", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_sample_size.tiff")))
# knitr::include_graphics(here::here("figures/rmse_sample_size.png"))
```

```{r rmseAuc, echo=FALSE, warning=FALSE, message=FALSE}
selectedScenario <- analysisIds %>%
  dplyr::filter(
    base == "moderate",
    type == "constant",
    sampleSize == 4250,
    auc == 0.85,
    harm == "absent"
  ) %>%
  dplyr::pull(scenario)

rmseBaseCaseAuc <- rmse %>% dplyr::filter(scenarioId == selectedScenario)

selectedScenario <- analysisIds %>%
  dplyr::filter(
    base == "moderate",
    type == "linear-high",
    sampleSize == 4250,
    auc == 0.85,
    harm == "absent"
  ) %>%
  dplyr::pull(scenario)

rmseLinearInteractionAuc <- rmse %>% dplyr::filter(scenarioId == selectedScenario)

selectedScenario <- analysisIds %>%
  dplyr::filter(
    base == "moderate",
    type == "quadratic-high",
    sampleSize == 4250,
    auc == 0.75,
    harm == "absent"
  ) %>%
  dplyr::pull(scenario)

rmseQuadraticAuc <- rmse %>% dplyr::filter(scenarioId == selectedScenario)

selectedScenario <- analysisIds %>%
  dplyr::filter(
    base == "moderate",
    type == "non-monotonic",
    sampleSize == 4250,
    auc == 0.75,
    harm == "absent"
  ) %>%
  dplyr::pull(scenario)

rmseNonMonotonicAuc <- rmse %>% dplyr::filter(scenarioId == selectedScenario)
```

Similarly, when we increased the AUC of the true prediction model to 0.85 (OR =
0.8 and N = 4,250), RCS-3 had the lowest RMSE in the case of strong quadratic or
non-monotonic deviations and very comparable performance to the – optimal –
linear interaction model in the case of strong linear deviations (median RMSE of
`r median(rmseLinearInteractionAuc$rcs_3_knots)` for RCS-3 compared to 
`r median(rmseLinearInteractionAuc$linear_predictor)` for the linear interaction
model). Similar to the base case scenario the adaptive approach wrongly selected
the constant treatment effect model
(`r adaptiveSelections %>% filter(scenarioId == 297) %>% pull(treatment) * 100`\%
and
`r adaptiveSelections %>% filter(scenarioId == 405) %>% pull(treatment) * 100`\%
of the replications in the
strong linear and non-monotonic deviation scenarios without treatment-related
harms, respectively), leading to increased variability of the RMSE (Supplement,
Figure S5).

```{r rmseauc, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in simulated samples 4,250. True prediction AUC of 0.85. RMSE was calculated on a super-population of size 500,000", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_moderate_auc.tiff")))
# knitr::include_graphics(here::here("figures/rmse_auc.png"))
```

With a true constant relative treatment effect, discrimination for benefit was
only slightly lower for the linear interaction model, but substantially lower
for the non-linear RCS approaches (Figure \ref{fig:discrimination}; panel
A). With strong linear or quadratic deviations from a constant relative
treatment effect, all methods discriminated quite similarly (Figure
\ref{fig:discrimination}; panels B-C). With non-monotonic deviations, the
constant effect model had much lower discriminative ability compared to all
other methods
(median AUC of 0.500 for the constant effects model,
`r aucDistribution %>% filter(scenarioId == 397) %>% pull(median_linear)`
for the linear interaction model and
`r aucDistribution %>% filter(scenarioId == 397) %>% pull(median_rcs_3)`
Figure \ref{fig:discrimination}; panel D).
The adaptive approach was unstable in terms of discrimination for benefit,
especially with treatment-related harms. With increasing number of RCS knots, we
observed decreasing median values and increasing variability of the
c-for-benefit in all scenarios. When we increased the sample size to 17,000 we
observed similar trends, however the performance of all methods was more stable
(Supplement, Figure S6). Finally, when we increased the true prediction AUC to
0.85 the adaptive approach was, again, more conservative, especially with
non-monotonic deviations and null or moderate treatment-related harms
(Supplement, Figure S5).

```{r discrimination, cache=TRUE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated samples of size 4,250. True prediction AUC of 0.75.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_base.tiff")))
# knitr::include_graphics(here::here("figures/discrimination_base.png"))
```

In terms of calibration for benefit, the constant effects model outperformed all
other models in the scenario with true constant treatment effects, but was
miscalibrated for all deviation scenarios (Figure \ref{fig:calibration}). The
linear interaction model showed best or close to best calibration across all
scenarios and was only outperformed by RCS-3 in the case of non-monotonic
deviations and treatment-related harms (Figure \ref{fig:calibration}; panel
D). The adaptive approach was worse calibrated under strong linear and
non-monotonic deviations compared to the linear interaction model and
RCS-3. When we increased the sample size to 17,000 (Supplement, Figure S6) or
the true prediction AUC to 0.85 (Supplement, Figure S7), RCS-3 was somewhat
better calibrated than the linear interaction model with strong quadratic
deviations.

```{r calibration, cache=TRUE, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.75 and sample size of 4,250.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/calibration_moderate_base.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

The results from all individual scenarios can be explored online at
[https://arekkas.shinyapps.io/simulation_viewer/](https://arekkas.shinyapps.io/simulation_viewer/). Additionally,
all the code for the simulations can be found at
[https://github.com/rekkasa/arekkas_HteSimulation_XXXX_2021](https://github.com/rekkasa/arekkas_HteSimulation_XXXX_2021)

## Empirical illustration

```{r, echo=FALSE, warning=FALSE, message=FALSE}
gustoPerformance <- readr::read_csv(here::here("data/processed/gustoPerformanceMetrics.csv"))

discriminationPerformance <- gustoPerformance %>%
  select("discrimination")
calibrationPerformance <- gustoPerformance %>%
  select("calibration")
```

We used the derived prognostic index to fit a constant treatment effect, a
linear interaction and an RCS-3 model individualizing absolute benefit
predictions. Following our simulation results, RCS-4 and RCS-5 models were
excluded. Finally, an adaptive approach with the 3 candidate models was applied.

All considered methods provided similar fits, predicting increasing benefits for
patients with higher baseline risk predictions, and followed the evolution of
the stratified estimates closely (Figure \ref{fig:gusto}). The constant
treatment effect model had somewhat lower AIC compared to the linear interaction
model (AIC: 
`r gustoPerformance %>% filter(method == "constant treatment effect") %>% pull(aic) %>% round()` versus 
`r gustoPerformance %>% filter(method == "RCS smoothing with 3 knots") %>% pull(aic) %>% round()`),
equal cross-validated discrimination
(c-for-benefit:
`r gustoPerformance %>% filter(method == "constant treatment effect") %>% pull(discrimination)`),
and slightly better cross-validated calibration
(ICI-for benefit:
`r gustoPerformance %>% filter(method == "constant treatment effect") %>% pull(calibration)` versus 
`r gustoPerformance %>% filter(method == "RCS smoothing with 3 knots") %>% pull(calibration)`).
In conclusion, although the sample size (`r nrow(gusto)` patients; 
`r sum(gusto$outcome)` events) allowed for flexible modeling approaches, a 
simpler constant treatment effect model is adequate for predicting absolute
30-day mortality benefits of treatment with tPA in patients with acute MI.

```{r gusto, cache=TRUE, echo=FALSE, fig.cap="Individualized absolute benefit predictions based on baseline risk when using a constant treatment effect approach, a linear interaction approach and RCS smoothing using 3 knots. Risk stratified estimates of absolute benefit are presented within quartiles of baseline risk as reference.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/gusto.tiff")))
# knitr::include_graphics(here::here("figures/gusto.png"))
```

# Discussion

The linear interaction and the RCS-3 models displayed very good performance
under many of the considered simulation scenarios. The linear interaction model
was optimal in cases with moderate sample sizes (4.250 patients; ~785 events)
and moderately performing baseline risk prediction models, that is, it had lower
RMSE, was better calibrated for benefit and had better discrimination for
benefit, even in scenarios with strong quadratic deviations. In scenarios with
true non-monotonic deviations, the linear interaction model was outperformed by
RCS-3, especially in the presence of treatment-related harms. Increasing the
sample size or the prediction model’s discriminative ability favored RCS-3,
especially in scenarios with strong non-linear deviations from a constant
treatment effect.

Our simulation results clearly express the trade-off between the advantages of
flexibly modeling the relationship between baseline risk and treatment effect
and the disadvantages of overfitting this relationship to the sample at hand. With
infinite sample size, the more flexible approach (here RCS) will be optimal, but
in practice, with limited sample size, parsimonious models may be
preferable. Even with the substantial sample size of our base case scenario, the
(less flexible) linear interaction model performed better than the (more
flexible) RCS approach for most simulation settings. The even less flexible
constant treatment effect model, however, was only optimal when the treatment
effect was truly constant. Moreover, the assumption of a constant treatment
effect may often be too strong [@Rothwell1995; @Kent2016].  For example, infants
at lower risk of bronchopulmonary dysplasia benefit relatively more from vitamin
A therapy than infants at higher risk [@Rysavy2021]; higher risk prediabetic
patients benefit relatively more from metformin than lower risk patients
[@Sussman2015]. Hence, a linear interaction between baseline risk and the effect
of treatment may be the most sensible approach with moderate sample sizes.

RCS-4 and RCS-5 were too flexible in all considered scenarios, as indicated by
higher RMSE, increased variability of discrimination for benefit and worse
calibration of benefit predictions. Even with larger sample sizes and strong
quadratic or non-monotonic deviations, these more flexible methods did not
outperform the simpler RCS-3 approach. Higher flexibility may only be helpful
under more extreme patterns of HTE compared to the quadratic deviations
considered here. Considering interactions in RCS-3 models as the most complex
approach often may be reasonable.

Increasing the discriminative ability of the risk model reduced RMSE for all
methods. Higher discrimination translates in higher variability of predicted
risks, which, in turn, allows the considered methods to better capture absolute
treatment benefits. As a consequence, better risk discrimination also led to
higher discrimination between those with low or high benefit (as reflected in
values of c-for-benefit).

The adaptive approach had adequate median performance, following the “true”
model in most scenarios. With smaller sample sizes it tended to miss the
treatment-baseline risk interaction and selected simpler models (Supplement
Section 4). This conservative behavior resulted in increased RMSE variability in
these scenarios, especially with true strong linear or non-monotonic
deviations. Therefore, with smaller sample sizes the simpler linear interaction
model may be a safer choice for predicting absolute benefits, especially in the
presence of any suspected treatment-related harms.

One limitation is that we assumed treatment benefit to be a function of baseline
risk in the majority of the simulation scenarios. We attempted to expand our
scenarios by considering constant moderate and strong treatment-related harms,
applied on the absolute scale, in line with previous work [@Glasziou1995]. In a
limited set of scenarios with true interactions between treatment assignment and
covariates, our conclusions remained unchanged (Supplement, Section 7). Even
though the average error rates increased for all the considered methods, due to
the miss-specification of the outcome model, the linear interaction model had
the lowest error rates. RCS-3 had very comparable performance. The constant
treatment effect model was often biased, especially with moderate or strong
treatment-related harms. Future simulation studies could explore the effect of
more extensive deviations from risk-based treatment effects.

We only focused on risk-based methods, using baseline risk as a reference in a
two-stage approach to individualizing benefit predictions. However, there is a
plethora of different methods, ranging from treatment effect modeling to
tree-based approaches available in more recent literature [@Athey2019; @Lu2018;
@Wager2018; @powers2018some]. Many of these methods rely on incorporating
treatment-covariate interactions when predicting benefit. An important caveat of
such approaches is their sensitivity to overfitting, which may exaggerate the
magnitude of predicted benefits. In a wide range of simulation settings, a
simpler risk modeling approach was consistently better calibrated for benefit
compared to more complex treatment effect modelling approaches
[@vanKlaveren2019]. Similarly, when SYNTAX score II, a model developed for
identifying patients with complex coronary artery disease that benefit more from
percutaneous coronary intervention or from coronary artery bypass grafting was
redeveloped using fewer treatment-covariate interactions had better external
performance compared to its predecessor [@farooq2013anatomical;
@takahashi2020redevelopment]. However, whether this remains the case in a range
of empirical settings still needs to be explored.

In conclusion, the linear interaction approach is a viable option with moderate
sample sizes and/or moderately performing risk prediction models, assuming a
non-constant relative treatment effect plausible. RCS-3 is a better option with
more abundant sample size and when non-monotonic deviations from a constant
relative treatment effect and/or substantial treatment-related harms are
anticipated. Increasing the complexity of the RCS models by increasing the
number of knots does not improve benefit prediction. Using AIC for model
selection is attractive with larger sample size.

### 2022-04-12 extras/bookdown/supplement.Rmd
initialize

# Supplement

```{r, echo=FALSE, message=FALSE, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(here)
library(rms)

d <- function(x, decimals = 2) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}
knit_hooks$set(
  inline = function(x) {
    prettyNum(
      x,
      format = "f",
      big.mark = ",",
      decimal.mark = ".",
      preserve.width = "individual"
    )
  }
)
```

## Notation
We observe RCT data $(Z, X, Y)$, where for each patient $Z_i= 0, 1$ is the
treatment status, $Y_i = 0, 1$ is the observed outcome and $X_i$ is a set of
covariates measured. Let $\{Y_i(z), z=0, 1\}$ denote the unobservable potential
outcomes. We observe $Y_i = Z_iY_i(1) + (1 - Z_i)Y_i(0)$. We are interested in
predicting the conditional average treatment effect (CATE), 
$$\tau(x) = E\{Y(0) - Y(1)|X=x\}$$ 
Assuming that $(Z, X, Y)$ is a random sample from the target population and that
$\big(Y(0), Y(1)\big)\indep Z|X$, as we are in the RCT setting, we can predict
CATE from
\begin{align*}
\tau(x) &= E\{Y(0)\given X=x\}-E\{Y(1)\given X=x\}\\
&=E\{Y\given X=x, Z=0\}-E\{Y\given X=x, Z=1\}
\end{align*}

Based on an estimate of baseline risk
$$
E\{Y\given X=x, Z=0\}=g\big(\hat{lp}(x)\big)
$$
with $\hat{u}=\hat{lp}(x)=x^t\hat{\beta}$ the linear predictor and $g$ the link function,
we predict CATE from
$$
\hat{\tau}(x) = g\big(f(\hat{u}, 0)\big) - g\big(f(\hat{u}, 1)\big)
$$
where $f(u,z)$ describes interactions of the baseline risk linear predictor with
treatment.

## Simulation settings
For all patients we observe covariates $X_1,\dots,X_8$, of which $4$ are
continuous and $4$ are binary. More specifically,

\begin{equation*}
X_1,\dots,X_4 \sim N(0, 1)
\end{equation*}
\begin{equation*}
X_5,\dots,X_8 \sim B(1, 0.2)
\end{equation*}

We first, generate the binary outcomes $Y$ for the untreated patients ($Z=0$),
based on

\begin{equation} 
P(Y=1\given X=x, Z=0) = g(\beta_0 + \beta_1x_1+\dots+\beta_8x_8) = g(lp_0),
(\#eq:p0)
\end{equation}

where $$g(x) = \frac{e^x}{1+e^x}$$

For treated patients, outcomes are generated from:

\begin{equation}
P(Y = 1\given X=x, Z=1) = g(lp_1)
(\#eq:p1)
\end{equation}

where $$lp_1 = \gamma_2(lp_0-c)^2+\gamma_1(lp_0-c)+\gamma_0$$

## Base-case scenario
The base-case scenario assumes a constant odds ratio of $0.8$ in favor of
treatment. The simulated datasets are of size $n=4250$, where treatment is
allocated at random using a 50/50 split (80% power for the detection of an
unadjusted OR of 0.8, assuming an event rate of 20% in the untreated
arm). Outcome incidence in the untreated population is set at $20\%$. For the
development of the prediction model we use the model defined in \@ref(eq:p0)
including a constant treatment effect. When doing predictions, $Z$ is set to
$0$. The value of the true $\beta$ is such that the above prediction model
has an AUC of $0.75$.

The previously defined targets are achieved when $\beta=(-2.08,
0.49,\dots,0.49)^t$. For the derivations in the treatment arm we use
$\gamma=(\log(0.8), 1, 0)^t$.

## Deviations from base-case
We deviate from the base-case scenario in two ways. First, we alter the overall
target settings of sample size, overall treatment effect and prediction model
AUC. In a second stage, we consider settings that violate the assumption of a
constant relative treatment effect, using a model-based approach.

For the first part, we consider:

* Sample size:
  + $n=1064$
  + $n=17000$
* Overall treatment effect:
  + $OR=0.5$
  + $OR=1$
* Prediction performance:
  + $AUC=0.65$
  + $AUC=0.85$

We set the true risk model coefficients to be
$\bm{\beta} = \big(-1.63,0.26,\dots,0.26\big)^t$ for $AUC=0.65$ and $\bm{\beta} = \big(-2.7,0.82,\dots,0.82\big)^t$ 
for $AUC=0.85$. In both cases, $\beta_0$ is selected so that an event rate of
$20\%$ is maintained in the control arm.

For the second part linear, quadratic and non-monotonic deviations from the
assumption of constant relative effect are considered. We also consider
different intensity levels of these deviations. Finally, constant absolute
treatment-related harms are introduced, i.e. positive 
($0.25\times\text{true average benefit}$),
strong positive ($0.50\times\text{true average benefit}$) and negative
($-0.25\times\text{true average benefit}$; i.e. constant absolute
treatment-related benefit). In case of true absent treatment effects,
treatment-related harms are set to $1\%, 2\%$ and $-1\%$ for positive, strong
positive and negative setting, respectively. The settings for these deviations
are defined in Table \@ref(tab:tab1).

```{r tab1, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
asd <- read_csv(
  here("data/processed/analysisIds.csv")
)
kableExtra::kbl(
  asd,
  format = "html",
  escape = FALSE,
  longtable   = TRUE,
  align       = "r",
  booktabs    = TRUE,
  caption     = "Scenario settings of the entire simulation study.",
  col.names   = c(
    "Scenario",
    "Base",
    "Type",
    "N",
    "AUC",
    "Treatment-related harm",
    paste0("b", 0:8),
    paste0("g", 0:2),
    "c",
    "Before harms",
    "After harms"
  ),
  linesep = c('', '', '', '\\addlinespace'),
  digits      = 3,
  format.args = list(
    big.mar      = ",",
    decimal.mark = "."
  )
) %>%
  kableExtra::landscape() %>%
  kableExtra::add_header_above(
    c(
      "Analysis ID"           = 6,
      "Baseline risk"         = 9,
      "True treatment effect" = 4,
      "Benefit"               = 2
      )
  ) %>%
  kableExtra::kable_styling(font_size = 9) %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```

\newpage
## Plausible scenario settings

In this section we present specific scenarios from our simulation settings in
which evolution of benefit followed similar patterns to [@Kent2016]. In this
case patients were stratified into risk quarters based on their true baseline
risk. Within each risk quarter we constructed boxplots of true benefit.

## Approaches to individualize benefit predictions

### Risk modeling
Merging treatment arms, we develop prediction models including a constant relative treatment effect:

\begin{equation}
P(Y=1\given X=x,Z=z) = g(x^t\beta + \gamma z)
(\#eq:risk)
\end{equation}

We derive baseline risk predictions for patients by setting $Z=0$ in
\@ref(eq:risk). All methods for individualizing benefit predictions are 2-stage
methods, that start by fitting a model for predicting baseline risk. The
estimated linear predictor of this model is

\begin{equation*}
\hat{lp} = lp(x;\hat{\beta}) = x^t\hat{\beta}
\end{equation*}

### Risk stratification
Derive a prediction model using the same approach as above and divide the
population in equally sized risk-based subgroups. Estimate subgroup-specific
absolute benefit from the observed absolute differences. Subject-specific
benefit predictions are made by attributing to individuals their corresponding
subgroup-specific estimate.

### Constant treatment effect
Assuming a constant relative treatment effect, fit the adjusted model in
\@ref(eq:risk). Then, predict absolute benefit using

\begin{equation}
\hat{\tau}(x;\hat{\beta},\hat{\gamma})=g(f(\hat{lp}, 0)) - g(f(\hat{lp},1)), 
(\#eq:main)
\end{equation}

where $f(\hat{lp}, z) = \hat{lp}+\hat{\gamma}z$, with $\hat{\gamma}$ the
estimated relative treatment effect (log odds ratio).

### Linear interaction
We relax the assumption of a constant relative treatment effect in
\@ref(eq:main) by setting

$$ f(\hat{lp}, z) = \gamma_0+\gamma_1z+\gamma_2\hat{lp}+\gamma_3z\hat{lp} $$

### Restricted cubic splines
Finally, we drop the linearity assumption and predict absolute benefit using
smoothing with restricted cubic splines with $k=3, 4$ and $5$ knots. More
specifically, we set:

$$ f(\hat{lp}, z) = \gamma_0 + \gamma_1z+zs(\hat{lp}) $$
where
$$s(x)=\alpha_0+\alpha_1h_1(x)+\alpha_2h_2(x)+\dots+\alpha_{k-1}h_{k-1}(x)$$
with $h_1(x)=x$ and for $j=2,\dots,k-2$
$$h_{j+1}(x)= (x-t_j)^3-(x-t_{k-1})_+^3 \frac{t_k-t_j}{t_k-t_{k-1}}+(x-t_k)^3_+\frac{t_{k-1}-t_j}{t_k-t_{k-1}}$$
where 
$t_1,\dots,t_k$ are the selected knots [@harrell2017regression].

\newpage
## Adaptive model selection frequencies

```{r adaptiveBase, cache=TRUE, echo=FALSE, fig.cap="Model selection frequencies of the adaptive approach based on Akaike's Information Criterion across 500 replications. The scenario with the true constant relative treatment effect (first panel) had a true prediction AUC of 0.75 and sample size of 4,250. ", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/selected_model_adaptive_base.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

```{r adaptiveSampleSize, cache=TRUE, echo=FALSE, fig.cap="Model selection frequencies of the adaptive approach based on Akaike's Information Criterion across 500 replications. Sample size is 17,000 rather than 4,250 in Figure \\ref{fig:adaptiveBase}", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/selected_model_adaptive_sample_size.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

```{r adaptiveAuc, cache=TRUE, echo=FALSE, fig.cap="Model selection frequencies of the adaptive approach based on Akaike's Information Criterion across 500 replications. AUC is 0.85 rather than 0.75 in Figure \\ref{fig:adaptiveBase}", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/selected_model_adaptive_auc.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

\newpage
## Discrimination and calibration for benefit
The c-for-benefit represents the probability that from two randomly chosen
matched patient pairs with unequal observed benefit, the pair with greater
observed benefit also has a higher predicted benefit. To be able to calculate
observed benefit, patients in each treatment arm are ranked based on their
predicted benefit and then matched 1:1 across treatment arms. Observed treatment
benefit is defined as the difference of observed outcomes between the untreated
and the treated patient of each matched patient pair. Predicted benefit is
defined as the average of predicted benefit within each matched patient pair.

We evaluated calibration in a similar manner, using the integrated calibration
index (ICI) for benefit [@Austin2019]. The observed benefits are regressed on the
predicted benefits using a locally weighted scatterplot smoother (loess). The
ICI-for-benefit is the average absolute difference between predicted and smooth
observed benefit. Values closer to represent better calibration.

```{r discriminationSampleSize, cache=TRUE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.75 and sample size of 17,000", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_sample_size.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

```{r discriminationAuc, cache=TRUE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.85 and sample size of 4,250", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/discrimination_moderate_auc.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

```{r calibrationSampleSize, cache=TRUE, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.75 and sample size of 17,000", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/calibration_moderate_sample_size.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

```{r calibrationAuc, cache=TRUE, echo=FALSE, fig.cap="Calibration for benefit of the considered methods across 500 replications calculated in a simulated sample of size 500,000. True prediction AUC of 0.85 and sample size of 4,250", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/calibration_moderate_auc.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

\newpage
## Strong relative treatment effect

Here we present the root mean squared error of the considered methods using
strong constant relative treatment effect ($\text{OR}=0.5$) as the
reference. Again, the same sample size and prediction performance settings were
considered along with the same settings for linear, quadratic and non-monotonic
deviations from the base case scenario of constant relative treatment effects
are considered. All results can be found at
[https://arekkas.shinyapps.io/simulation_viewer/](https://arekkas.shinyapps.io/simulation_viewer/).

```{r rmseHighBase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated super-population of size 500,000. The scenario with true constant relative treatment effect (panel A) had a true prediction AUC of 0.75 and sample size of 4,250. The RMSE is also presented for strong linear (panel B), strong quadratic (panel C), and non-monotonic (panel D) deviations from constant relative treatment effects. Panels on the right side present the true relationship between baseline risk (x-axis) and absolute treatment benefit (y-axis). The 2.5, 25, 75 and 97.5 percentiles of the risk distribution are expressed in the boxplot.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_high_base.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

```{r rmseHighSampleSize, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replcations calculated in a simulated sample of size 500,000. Sample size is 17,000 rather than 4,250 in Figure \\ref{fig:rmseHighBase}.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_high_sample_size.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

```{r rmseHighAuc, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replcations calculated in a simulated sample of size 500,000. AUC is 0.85 rather than in Figure \\ref{fig:rmseHighBase}.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_high_auc.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

\newpage
## Treatment interactions

We carried out a smaller set of simulations, in which we assumed true
treatment-covariate interactions. Sample size was set to 4,250 and the AUC of
the true prediction model was set to 0.75. The following scenarios were
considered: 1) 4 true weak positive interactions
($\text{OR}_{Z=1} / \text{OR}_{Z=0}=0.83$); 2) 4 strong positive
interactions ($\text{OR}_{Z=1} / \text{OR}_{Z=0}=0.61$); 3) 2 weak and 2
strong positive interactions; 4) 4 weak negative interactions 
($\text{OR}_{Z=1} / \text{OR}_{Z=0}=1.17$); 5) 4 strong negative interactions 
($\text{OR}_{Z=1} / \text{OR}_{Z=0}=1.39$); 6) 2 weak and 2 strong negative
interactions; 7) combined positive and negative strong interactions. We also
considered constant treatment-related harms applied on the absolute scale to all
treated patients. The exact settings were: 1) absent treatment-related harms; 2)
moderate treatment-related harms, defined as 25\% of the average true benefit of the
scenario without treatment-related harms; 3) strong treatment-related harms
defined as 50\% of the true average benefit of the scenario without
treatment-related harms; 4) negative treatment-related harms (benefit), defined
as an absolute risk reduction for treated patients of 50\% of the true average
benefit of the scenario without treatment-related harms. The exact settings can
be found in Table \@ref(tab:tab2).

```{r tab2, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
asd <- read_csv(
  here("data/processed/analysisIdsInteractions.csv")
)
kableExtra::kbl(
  asd,
  format = "html",
  escape = FALSE,
  longtable = TRUE,
  align = "r",
  booktabs = TRUE,
  caption = "Scenario settings of the treatment-covariate interaction scenarios.",
  col.names = c(
    "Scenario",
    "Base",
    "Type",
    "N",
    "AUC",
    "Treatment-related harm",
    paste0("b", 0:8),
    "g0", "g1", "g2", "g5", "g6",
    "Before harms",
    "After harms"
  ),
  linesep = c('', '', '', '\\addlinespace'),
  digits      = 2,
  format.args = list(
    big.mar      = ",",
    decimal.mark = "."
  )
) %>%
  kableExtra::landscape() %>%
  kableExtra::add_header_above(
    c(
      "Analysis ID"           = 6,
      "Baseline risk"         = 9,
      "True treatment effect" = 5,
      "Benefit"               = 2
      )
  ) %>%
  kableExtra::kable_styling(font_size = 9) %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```

```{r rmseInteractionPositive, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000 where treatment-covariate interactions all favoring treatment were considered.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction_positive.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```
```{r rmseInteractionNegative, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000 where treatment-covariate interactions all favoring the control were considered.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction_negative.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```
```{r rmseInteractionCombined, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated in a simulated sample of size 500,000 where treatment-covariate interactions 2 favoring treatment and 2 favoring the control were considered.", fig.show="hold", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/rmse_interaction_combined.tiff")))
# knitr::include_graphics(here::here("figures/calibration_base.png"))
```

\newpage
## Empirical illustration

For predicting baseline risk of 30-day mortality we fitted a logistic regression
model with age, Killip class (*Killip*), systolic blood pressure (*sysbp*),
pulse rate (*pulse*), prior myocardial infarction (*pmi*), location of
myocardial infarction (*miloc*) and treatment as the covariates. Baseline
predictions were made setting treatment to 0.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
load(here::here("data/raw/gusto.rda"))
gusto <- gusto %>%
  tibble() %>%
  filter(!is.na(tpa))

gusto <- gusto %>%
  tibble() %>%
  filter(tx != "SK+tPA") %>%
  rename(
    "outcome" = "day30",
    "treatment" = "tpa"
  )

treatmentArms <- gusto %>%
  group_by(treatment) %>%
  summarise(n = n())

prediction <- glm(
   outcome ~ treatment + age + Killip + pmin(sysbp, 120) + lsp(pulse, 50) + pmi + miloc,
   data = gusto,
   family = "binomial",
   maxit = 99
)

predictionCoefficients <- coef(prediction)
predictionCoefficients <- predictionCoefficients[c(1, 3:12, 2)]
```

\begin{equation*}
P(outcome=1|X=x) = \text{expit}(lp(x)),
(\#eq:gusto1)
\end{equation*}
where 
\begin{equation*}
\begin{aligned}
lp(x)=& \beta_0 + \beta_1 \text{age} + \beta_2 I(\text{Killip}=II) + \beta_3I(\text{Killip}=III) +\\
&\beta_4 I(\text{Killip}=IV) + \beta_5min(\text{sysbp}, 120) + \beta_6 \text{pulse}+\\
&\beta_7 max(\text{pulse - 50, 0}) + \beta_8 I(\text{pmi}=yes)+\\
&\beta_9 I(\text{miloc}=Anterior) + \beta_9 I(\text{miloc}=Other) +\\
&\gamma\times\text{treatment}
\end{aligned}
(\#eq:gusto2)
\end{equation*}

and $expit(x) = \frac{e^x}{1+e^x}$

```{r, echo=FALSE}
predictionSummary <- summary(prediction)
# predictionSummaryCoef <- tibble(predictionSummary$coefficients) %>%
#   rownames_to_column("Variable")

predictionCoefficientsTable <- predictionSummary$coefficients %>%
  as.data.frame() %>%
  mutate(
    rownames = c(
      "Intercept",
      "Age",
      "Killip class = II",
      "Killip class = III",
      "Killip class = IV",
      "Systolic blood pressure",
      "Pulse rate (1)",
      "Pulse rate (2)",
      "Previous MI (yes)",
      "MI location (Other)",
      "MI location (Anterior)",
      "Treatment"
    )
  ) %>%
  rownames_to_column("Variable") %>%
  select(-Variable) %>%
  select(rownames, everything())

colnames(predictionCoefficientsTable) <- c(
  "Variable",
  "Estimate",
  "stderror",
  "zvalue",
  "pvalue"
)

kableExtra::kbl(
  predictionCoefficientsTable,
  format = "latex",
  escape = FALSE,
  longtable = TRUE,
  align = c("l", rep("r", 4)),
  booktabs = TRUE,
  caption = "Coefficients of the prediction model for 30-day mortality, based on the data from GUSTO-I trial.",
  digits = 3,
  linesep = "",
  format.args = list(
    big.mar      = ",",
    decimal.mark = "."
  )
) %>%
  kableExtra::kable_styling(font_size = 7)
```

### 2022-04-12 submission/manuscript.rmd
initialize

--- 2022-04-12 extras/bookdown/index.Rmd
@@
 ---
-author: Alexandros Rekkas
-date: "`r Sys.Date()`"
-site: bookdown::bookdown_site
-documentclass: book
-biblio-style: apalike
-bibliography: [ "references.bib" ]
-link-citations: true
+title: |
+ Individualized treatment effect was predicted best by modeling 
+ baseline risk in interaction with treatment assignment
+abstract: |
+  **Objective**: To compare different risk-based methods for optimal prediction
+  of treatment effects. **Study Design and Setting**: We simulated RCT data
+  using diverse assumptions for the average treatment effect, a baseline
+  prognostic index of risk (PI), the shape of its interaction with treatment
+  (none, linear, quadratic or non-monotonic), and the magnitude of
+  treatment-related harms (none or constant independent of the PI). We predicted
+  absolute benefit using: models with a constant relative treatment effect;
+  stratification in quarters of the PI; models including a linear interaction of
+  treatment with the PI; models including an interaction of treatment with a
+  restricted cubic spline (RCS) transformation of the PI; an adaptive approach
+  using Akaike’s Information Criterion. We evaluated predictive performance
+  using root mean squared error and measures of discrimination and calibration
+  for benefit. **Results**: The linear-interaction model displayed optimal or
+  close-to-optimal performance across many simulation scenarios with moderate
+  sample size (N=4,250; ~800 events). The RCS-model was optimal for strong
+  non-linear deviations from a constant treatment effect, particularly when
+  sample size was larger (N=17,000). The adaptive approach also required larger
+  sample sizes. These findings were illustrated in the GUSTO-I
+  trial. **Conclusion**: An interaction between baseline risk and treatment
+  assignment should be considered to improve treatment effect predictions.
+author:
+  - name: Alexandros Rekkas
+    affiliation: 1
+    correspondingauthor: true
+  - name: Peter R. Rijnbeek
+    affiliation: 1
+  - name: David M. Kent
+    affiliation: 2
+  - name: Ewout W. Steyerberg
+    affiliation: 3
+  - name: David van Klaveren
+    affiliation: 4
+address:
+  - code: 1
+    address: Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
+  - code: 2
+    address: Predictive Analytics and Comparative Effectiveness Center, Institute for Clinical Research and Health Policy Studies, Tufts Medical Center, Boston, Massachusetts, USA
+  - code: 3
+    address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
+  - code: 4
+    address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
+numbersections: true
+linenumbers: false
+journal: arXiv.org
+keywords:
+  - treatment effect heterogeneity
+  - absolute benefit
+  - prediction models
+output: 
+  rticles::elsevier_article:
+    keep_tex: true
+  bookdown::word_document2:
+    reference_docx: reference.docx
+geometry: margin=1.0in
+date: false
+toc: false
+font-size: 11pt
+header-includes:
+  - \renewcommand*\familydefault{\sfdefault}
+  - \usepackage{setspace}
+  - \usepackage{amsmath}
+  - \doublespacing
+  - \usepackage{amssymb}
+  - \usepackage{bm}
+  - \usepackage{caption}
+  - \usepackage{booktabs}
+  - \date{}
+  - \newcommand\given[1][]{\:#1\vert\:}
+  - \newcommand{\indep}{\perp \!\!\! \perp}
+editor_options: 
+  chunk_output_type: console
+bibliography: references.bib
 csl: jce.csl
 ---
-
-% Pandoc header
-\renewcommand*\familydefault{\sfdefault}
-\usepackage{setspace}
-\usepackage{amsmath}
-\doublespacing
-\usepackage{amssymb}
-\usepackage{caption}
-\usepackage{booktabs}
-\date{}
-\newcommand\given[1][]{\:#1\vert\:}
-\newcommand{\indep}{\perp \!\!\! \perp}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
 
@@
 
-$$\text{RMSE}=\sqrt{\frac{1}{n}\sum_{i=1}^n\big(\tau(x_i) - \hat{\tau}(x_i)\big)^2}$$
+$$\text{RMSE}=\sqrt{\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2}$$
 
@@
 selection is attractive with larger sample size.
+
+\newpage
+# References
+\setlength{\parindent}{-0.25in}
+\setlength{\leftskip}{0.25in}
+\noindent
+<div id="refs"></div>
+\setlength{\parindent}{0in}
+\setlength{\leftskip}{0in}
+\noindent

### 2022-04-12 submission/supplement.rmd
initialize

--- 2022-04-12 extras/bookdown/supplement.Rmd
@@
-# Supplement
+---
+title: "Supplementary material"
+output:
+    bookdown::pdf_document2: default
+geometry: margin=1.0in
+toc: true
+font-size: 12pt
+header-includes:
+  - \usepackage{amssymb}
+  - \usepackage{bm}
+  - \usepackage{booktabs}
+  - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
+  - \newcommand\given[1][]{\:#1\vert\:}
+  - \newcommand{\indep}{\perp \!\!\! \perp}
+editor_options: 
+  chunk_output_type: console
+bibliography: references.bib
+csl: jce.csl
+---
 
@@
 
-## Notation
+\beginsupplement
+\newpage
+
+# Notation
 We observe RCT data $(Z, X, Y)$, where for each patient $Z_i= 0, 1$ is the
@@
 
-## Simulation settings
+# Simulation settings
 For all patients we observe covariates $X_1,\dots,X_8$, of which $4$ are
@@
   here("data/processed/analysisIds.csv")
-)
+) %>%
+  select(-"type")
 kableExtra::kbl(
   asd,
-  format = "html",
+  format = "latex",
   escape = FALSE,
@@
     "Base",
-    "Type",
     "N",
@@
     c(
-      "Analysis ID"           = 6,
+      "Analysis ID"           = 5,
       "Baseline risk"         = 9,
@@
   ) %>%
-  kableExtra::kable_styling(font_size = 9) %>%
-  kableExtra::scroll_box(width = "100%", height = "400px")
+  kableExtra::kable_styling(font_size = 7)
 ```
@@
 \newpage
-## Plausible scenario settings
+# Plausible scenario settings
 
@@
 
-## Approaches to individualize benefit predictions
+```{r, cache=TRUE, echo=FALSE, warning=FALSE, fig.cap="Simulation scenarios that closely follow trials.
+In this case, we see increasing absolute benefits with increasing baseline risk.", fig.show="hold", out.width = '50%'}
+include_graphics(
+  c(
+    here("figures/scenario_251.png"),
+    here("figures/scenario_324.png")
+  )
+)
+```
 
-### Risk modeling
+```{r, cache=TRUE, echo=FALSE, warning = FALSE, fig.cap="Simulation scenarios that closely follow trials.
+In this case, we see increasing absolute benefits with increasing baseline risk up to the third risk quarter.
+In the fourth risk quarter this trend is interrupted and benefits are diminished.", fig.show="hold", out.width = '50%'}
+include_graphics(
+  c(
+    here("figures/scenario_406.png"),
+    here("figures/scenario_422.png")
+  )
+)
+```
+
+# Approaches to individualize benefit predictions
+
+## Risk modeling
 Merging treatment arms, we develop prediction models including a constant relative treatment effect:
@@
 P(Y=1\given X=x,Z=z) = g(x^t\beta + \gamma z)
+\end{equation}
 (\#eq:risk)
-\end{equation}
 
@@
 
-### Risk stratification
+## Risk stratification
 Derive a prediction model using the same approach as above and divide the
@@
 
-### Constant treatment effect
+## Constant treatment effect
 Assuming a constant relative treatment effect, fit the adjusted model in
@@
 
-### Linear interaction
+## Linear interaction
 We relax the assumption of a constant relative treatment effect in
@@
 
-### Restricted cubic splines
+## Restricted cubic splines
 Finally, we drop the linearity assumption and predict absolute benefit using
@@
 \newpage
-## Adaptive model selection frequencies
+# Adaptive model selection frequencies
 
@@
 \newpage
-## Discrimination and calibration for benefit
+# Discrimination and calibration for benefit
 The c-for-benefit represents the probability that from two randomly chosen
@@
 \newpage
-## Strong relative treatment effect
+# Strong relative treatment effect
 
@@
 \newpage
-## Treatment interactions
+# Treatment interactions
 
@@
   asd,
-  format = "html",
+  format = "latex",
   escape = FALSE,
@@
   ) %>%
-  kableExtra::kable_styling(font_size = 9) %>%
-  kableExtra::scroll_box(width = "100%", height = "400px")
+  kableExtra::kable_styling(font_size = 6)
 ```
@@
 \newpage
-## Empirical illustration
+# Empirical illustration
 
@@
 ```
+
+\newpage
+# References
+\setlength{\parindent}{-0.25in}
+\setlength{\leftskip}{0.25in}
+\noindent
+<div id="refs"></div>
+\setlength{\parindent}{0in}
+\setlength{\leftskip}{0in}
+\noindent

### 2022-04-12 code/AucPlots.R

@@ -1,0 +1,66 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+
+source("code/helpers/CreateManuscriptPlots.R")
+
+metric <- "rmse"
+titles <- c(
+  "AUC = 0.65",
+  "AUC = 0.85"
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) 
+
+nonConstantIds <- c(56, 57)
+
+for (i in seq_along(nonConstantIds)) {
+  tmp <- processed %>%
+    dplyr::filter(scenarioId == nonConstantIds[i])
+  
+  plotName <- paste0(
+    paste(metric, nonConstantIds[i], sep = "_"),
+    ".png"
+  )
+  
+  plot <- createPlot(
+    data = tmp,
+    metric = metric,
+    title = titles[i],
+    limits = c(0, .1)
+  )
+  
+  if (i > 1) {
+    plot <- plot +
+      ggplot2::theme(
+        legend.position = "none",
+        axis.title.x    = ggplot2::element_blank(),
+        axis.title.y    = ggplot2::element_blank(),
+        axis.text.x     = ggplot2::element_text(size = 16, angle = 45, hjust = 1),
+        axis.text.y     = ggplot2::element_text(size = 16),
+        plot.title      = ggplot2::element_text(size = 22)
+      )
+  } else {
+    plot <- plot +
+      ggplot2::theme(
+        legend.position = "none",
+        axis.title.x    = ggplot2::element_blank(),
+        axis.title.y    = ggplot2::element_text(size = 18),
+        axis.text.x     = ggplot2::element_text(size = 16, angle = 45, hjust = 1),
+        axis.text.y     = ggplot2::element_text(size = 16),
+        plot.title      = ggplot2::element_text(size = 22)
+      )
+  }
+  
+  
+    ggplot2::ggsave(file.path("figures", plotName), width = 8.5, height = 7)
+  
+}

### 2022-04-12 code/CalibrationBase.R

@@ -1,0 +1,168 @@
+#!/usr/bin/env Rscript
+
+# ===================================================
+# Description:
+#   Generates the calibration plots
+# Input:
+#   - sample size
+#   - auc
+#   - value
+# Output:
+#   - figures/calibration_*.tiff
+# Depends:
+#   - data/processed/analysisIds.csv
+#   - data/processed/calibration.csv
+#   - code/helpers/CreateManuscriptPlots.R
+#   - code/helpers/PlotResult.R
+# ===================================================
+
+args <- commandArgs(trailingOnly = TRUE)
+args_base <- as.character(args[1])
+args_sampleSize <- as.numeric(args[2])
+args_auc <- as.numeric(args[3])
+args_value <- as.character(args[4])
+
+library(tidyverse)
+library(glue)
+library(ggtext)
+source("code/helpers/CreateManuscriptPlots.R")
+source("code/helpers/PlotResult.R")
+
+scenarioIds <- readr::read_csv("data/processed/analysisIds.csv") %>%
+  filter(
+    base == args_base,
+    harm != "negative",
+    !(type %in% c("quadratic-moderate", "linear-moderate")),
+    sampleSize == args_sampleSize,
+    auc == args_auc
+  ) 
+
+metric    <- "calibration"
+
+titles <- scenarioIds %>%
+  mutate(
+    title = ifelse(
+      type == "constant",
+      str_to_sentence(glue("{str_replace_all(type, '-', ' ')} treatment effect")),
+      str_to_sentence(glue("{str_replace_all(type, '-', ' ')} deviation"))
+    )
+  ) %>%
+  select(title) %>%
+  unlist() %>%
+  unique()
+
+names(titles) <- NULL
+
+titlePrefix <- paste0(
+  "**",
+  LETTERS[1:4],
+  ".**"
+)
+
+titles <- paste(
+  titlePrefix,
+  titles
+)
+
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+f <- function(x) x * 100
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) %>% 
+  mutate_at(
+    c(
+      "constant_treatment_effect",
+      "stratified",
+      "linear_predictor",
+      "rcs_3_knots",
+      "rcs_4_knots",
+      "rcs_5_knots",
+      "adaptive"
+    ),
+    f
+  )
+scenarios <- scenarioIds %>%
+  filter(harm == "absent") %>%
+  select(scenario) %>%
[…]

### 2022-04-12 code/CalibrationInteractionPlots.R

@@ -1,0 +1,91 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+
+source("code/helpers/CreateManuscriptPlots.R")
+
+scenarios <- c(10, 64, 66, 65)
+
+metric <- "calibration"
+titles <- c(
+  "Base case scenario",
+  "Weak interactions",
+  "Mixed interactions",
+  "Strong interactions"
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+f <- function(x) x * 100
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) %>% 
+  mutate_at(
+    c(
+      "constant_treatment_effect",
+      "stratified",
+      "linear_predictor",
+      "rcs_3_knots",
+      "rcs_4_knots",
+      "rcs_5_knots",
+      "adaptive"
+    ),
+    f
+  )
+
+plotList <- list()
+
+for (i in seq_along(scenarios)) {
+  tmp <- processed %>%
+    dplyr::filter(scenarioId == scenarios[i])
+  
+  plot <- createPlot(
+    data = tmp,
+    metric = metric,
+    title = titles[i],
+    limits = c(0, 6),
+    pointSize = .5
+  ) +
+    ggplot2::theme(
+      legend.position = "none",
+      axis.title.x    = ggplot2::element_blank(),
+      axis.text.x     = ggplot2::element_text(size = 10, angle = 45, hjust = 1),
+      axis.title.y    = ggplot2::element_blank(),
+      axis.text.y     = ggplot2::element_text(size = 9.5),
+      plot.title      = ggtext::element_markdown(size = 14)
+    )
+  
+  plotList[[i]] <- plot
+}
+
+pp <- gridExtra::grid.arrange(
+  plotList[[1]],
+  plotList[[2]],
+  plotList[[3]],
+  plotList[[4]],
+  nrow = 2,
+  ncol = 2,
+  left = grid::textGrob(
+    expression(
+      paste(
+        "Integrated calibration index (x ", 
+        10^-2, 
+        ")"
+      )
+    ),
+    rot = 90
+  )
+)
+  
+  ggplot2::ggsave(
+    file.path("figures", "calibration_interactions.png"), 
+    plot = pp,
+    width = 11, 
+    height = 10.5
+  )
+  

### 2022-04-12 code/Clean.R

@@ -1,0 +1,4 @@
+#!/usr/bin/env Rscript
+
+renv::restore(rebuild = TRUE, clean = TRUE)
+tinytex::install_tinytex()

### 2022-04-12 code/CreateAdaptiveSelections.R

@@ -1,0 +1,39 @@
+#!/usr/bin/env Rscript
+
+# ===================================================
+# Description: 
+#   Creates a csv file with the models selected
+#   in each run of the simulation
+# Output: 
+#   - data/processed/adaptiveSelections.csv
+# Depends:
+#   - data/processed/adaptiveModel.csv
+# ===================================================
+
+library(tidyverse)
+
+readr::read_csv("data/processed/adaptiveModel.csv") %>%
+  group_by(scenarioId) %>% 
+  nest() %>% 
+  mutate(
+    counts = lapply(data, plyr::count),
+    n = lapply(data, nrow)
+  ) %>%
+  select(
+    scenarioId, counts, n
+  ) %>%
+  unnest(c(counts, n)) %>%
+  mutate(
+    perc = round(freq / n, 2),
+    model = case_when(
+      selectedAdaptiveModel == "RCS with 3 knots" ~ "rcs_3",
+      selectedAdaptiveModel == "RCS with 4 knots" ~ "rcs_4",
+      selectedAdaptiveModel == "RCS with 5 knots" ~ "rcs_5",
+      TRUE                                        ~ selectedAdaptiveModel
+    )
+  ) %>%
+  ungroup() %>%
+  select(scenarioId, model, perc) %>%
+  spread(model, perc) %>%
+  relocate(c(treatment, risk), .before = rcs_3) %>%
+  readr::write_csv("data/processed/adaptiveSelections.csv")

### 2022-04-12 code/CreateDistribution.R

@@ -1,0 +1,69 @@
+#!/usr/bin/env Rscript
+
+# =============================================================
+# Description:
+#   Creates a csv file with all median and the
+#   0.025 and 0.975 quantiles of the
+#   RMSE/AUC/calibration sampling distribution
+# Input:
+#   The metric we are interested in:
+#     - rmse
+#     - auc
+#     - calibration
+# Output:
+#   - data/processed/(rmse|discrimination|calibration)Distribution.csv
+# Depends:
+#   - code/helpers/GetMedians.R
+# =============================================================
+
+args <- commandArgs(trailingOnly = TRUE)
+args_metric <- as.character(args[1])
+saveFile <- paste0(
+  args_metric,
+  "Distribution.csv"
+)
+
+
+library(tidyverse)
+library(readr)
+
+source("code/helpers/GetMedians.R")
+
+read_csv(
+  file.path(
+    "data/processed",
+    paste0(args_metric, ".csv")
+  )
+) %>%
+  group_by(scenarioId) %>%
+  arrange(scenarioId) %>%
+  nest() %>%
+  mutate(
+    median = map(
+      data,
+      ~ getMedian(
+        data = .x
+      )
+    ),
+    lower = map(
+      data,
+      ~ getLower(
+        data = .x
+      )
+    ),
+    upper = map(
+      data,
+      ~ getUpper(
+        data = .x
+      )
+    )
+  ) %>%
+  ungroup() %>%
+  select(-data) %>%
+  unnest(c("median", "lower", "upper")) %>%
+  write_csv(
+    file.path(
+      "data/processed",
+      saveFile
+    )
+  )

### 2022-04-12 code/DiscriminationBase.R

@@ -1,0 +1,138 @@
+#!/usr/bin/env Rscript
+
+# ===================================================
+# Description:
+#   Generates the discrimination plots
+# Input:
+#   - sample size
+#   - auc
+#   - value
+# Output:
+#   - figures/rmse_*.tiff
+# Depends:
+#   - data/processed/analysisIds.csv
+#   - data/processed/discrimination.csv
+#   - code/helpers/CreateManuscriptPlots.R
+#   - code/helpers/PlotResult.R
+# ===================================================
+
+args <- commandArgs(trailingOnly = TRUE)
+args_base <- as.character(args[1])
+args_sampleSize <- as.numeric(args[2])
+args_auc <- as.numeric(args[3])
+args_value <- as.character(args[4])
+
+
+library(tidyverse)
+library(glue)
+library(ggtext)
+source("code/helpers/CreateManuscriptPlots.R")
+source("code/helpers/PlotResult.R")
+
+scenarioIds <- readr::read_csv("data/processed/analysisIds.csv") %>%
+  filter(
+    base == args_base,
+    harm != "negative",
+    !(type %in% c("quadratic-moderate", "linear-moderate")),
+    sampleSize == args_sampleSize,
+    auc == args_auc
+  ) 
+
+metric <- "discrimination"
+
+titles <- scenarioIds %>%
+  mutate(
+    title = ifelse(
+      type == "constant",
+      str_to_sentence(glue("{str_replace_all(type, '-', ' ')} treatment effect")),
+      str_to_sentence(glue("{str_replace_all(type, '-', ' ')} deviation"))
+    )
+  ) %>%
+  select(title) %>%
+  unlist() %>%
+  unique()
+
+names(titles) <- NULL
+
+titlePrefix <- paste0(
+  "**",
+  LETTERS[1:4],
+  ".**"
+)
+
+titles <- paste(
+  titlePrefix,
+  titles
+)
+
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+)
+scenarios <- scenarioIds %>%
+  filter(harm == "absent") %>%
+  select(scenario) %>%
+  unlist()
+names(scenarios) <- NULL
+
+plotList <- plotResult(scenarios, processed, titles, metric = metric, limits = c(.49, .6, .01))
+
+res <- gridExtra::grid.arrange(
+  plotList[[1]] +
+    theme(
+      plot.title = element_markdown(),
[…]

### 2022-04-12 code/DiscriminationInteractionPlots.R

@@ -1,0 +1,69 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+
+source("code/helpers/CreateManuscriptPlots.R")
+
+scenarios <- c(10, 64, 66, 65)
+
+metric <- "discrimination"
+titles <- c(
+  "Base case scenario",
+  "Weak interactions",
+  "Mixed interactions",
+  "Strong interactions"
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) 
+
+plotList <- list()
+
+for (i in seq_along(scenarios)) {
+  tmp <- processed %>%
+    dplyr::filter(scenarioId == scenarios[i])
+  
+  plot <- createPlot(
+    data = tmp,
+    metric = metric,
+    title = titles[i],
+    limits = c(.52, .61),
+    pointSize = .5
+  ) +
+    ggplot2::theme(
+      legend.position = "none",
+      axis.title.x    = ggplot2::element_blank(),
+      axis.text.x     = ggplot2::element_text(size = 9, angle = 45, hjust = 1),
+      axis.title.y    = ggplot2::element_blank(),
+      axis.text.y     = ggplot2::element_text(size = 9),
+      plot.title      = ggtext::element_markdown(size = 12)
+    )
+  
+  plotList[[i]] <- plot
+}
+
+
+pp <- gridExtra::grid.arrange(
+  plotList[[1]],
+  plotList[[2]],
+  plotList[[3]],
+  plotList[[4]],
+  nrow = 2,
+  ncol = 2,
+  left = "C-statistic for benefit"
+)
+  
+  ggplot2::ggsave(
+    file.path("figures", "discrimination_interactions.png"), 
+    plot = pp,
+    width = 11, 
+    height = 10.5,
+  )
+  

### 2022-04-12 code/GetGustoData.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+[[ ! -f "data/raw/gusto.rda" ]] && wget --output-document data/raw/gusto.rda -nc http://hbiostat.org/data/gusto.rda || echo -e "Nothing to be done"

### 2022-04-12 code/GustoPerformanceMetrics.R

@@ -1,0 +1,221 @@
+#!/usr/bin/env Rscript
+
+# Description:
+#   Generates the tables of c and ICI for benefit of all the methods used in the
+#   GUSTO-I application.
+# Output: 
+#   data/processed/gustoPerformanceMetrics.csv
+# Depends:
+#   data/raw/gusto.rda
+
+library(tidyverse)
+library(rms)
+library(SmoothHte)
+
+load("data/raw/gusto.rda")
+set.seed(19910930)
+
+gusto <- gusto %>%
+  tibble() %>%
+  rename(
+    "outcome" = "day30",
+    "treatment" = "tpa"
+  )
+
+replicationExpression <- function(gusto) {
+  prediction <- rms::lrm(
+    outcome ~ treatment + age + Killip + pmin(sysbp, 120) + lsp(pulse, 50) + pmi + miloc,
+    data = gusto,
+    maxit = 99
+  )
+  
+  riskLinearPredictor <- predict(
+    prediction,
+    newdata = gusto %>% mutate(treatment = 0) %>% data.frame()
+  )
+  
+  gusto <- gusto %>%
+    mutate(
+      riskLinearPredictor = riskLinearPredictor
+    )
+  
+  folds <- caret::createFolds(gusto$outcome, k = 4)
+  
+  evaluatePerformance <- function(data, testRows, settings) {
+    isStratified <- FALSE
+    if ("modelBased" %in% class(settings)) {
+      model <- fitModelBasedHte(data %>% dplyr::slice(-testRows), settings)
+      tmp <- data %>%
+        slice(testRows) %>%
+        dplyr::mutate(
+          predictedBenefit = predictBenefitModelBasedHte(
+            p             = plogis(.$riskLinearPredictor), 
+            modelBasedFit = model
+          )
+        )
+      
+    } else if ("rcs" %in% class(settings)) {
+      model <- fitRcsHte(data %>% dplyr::slice(-testRows), settings)
+      tmp <- gusto %>%
+        slice(testRows) %>%
+        dplyr::mutate(
+          predictedBenefit = predictSmoothBenefit(
+            p          = plogis(riskLinearPredictor), 
+            smoothFit  = model
+          )
+        )
+    } else if ("stratified" %in% class(settings)) {
+      isStratified <- TRUE
[…]

### 2022-04-12 code/GustoPlot.R

@@ -1,0 +1,158 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(rms)
+library(SmoothHte)
+
+load("data/raw/gusto.rda")
+maxRisk <- .3      # the maximum risk to be plotted
+
+gusto <- gusto %>%
+  tibble() %>%
+  filter(tx != "SK+tPA") %>%
+  rename(
+    "outcome" = "day30",
+    "treatment" = "tpa"
+  )
+
+prediction <- lrm(
+  outcome ~ treatment + age + Killip + pmin(sysbp, 120) + lsp(pulse, 50) + pmi + miloc,
+  data = gusto,
+  maxit = 99
+)
+
+riskLinearPredictor <- predict(
+  prediction,
+  newdata = gusto %>% mutate(treatment = 0) %>% data.frame()
+)
+
+gusto <- gusto %>%
+  mutate(
+    riskLinearPredictor = riskLinearPredictor
+  )
+
+constantModel <- fitModelBasedHte(
+  data     = gusto, 
+  settings = createModelBasedSettings(type = "treatment")
+)
+
+stratified <- fitStratifiedHte(
+  data     = gusto,
+  settings = createStratifiedSettings()
+)
+
+linearModel <- fitModelBasedHte(
+  data     = gusto,
+  settings = createModelBasedSettings()
+)
+
+rcs3Model <- fitRcsHte(
+  data     = gusto,
+  settings = createRcsSettings()
+)
+
+plot <- ggplot() +
+  geom_pointrange(
+    data = stratified$data,
+    aes(
+      x     = meanRisk, 
+      y     = estimate, 
+      ymin  = lower, 
+      ymax  = upper,
+      color = "Stratified"
+    ),
+    key_glyph = "rect"
+  ) +
+  stat_function(
+    data = data.frame(
+      x     = c(0, maxRisk), 
+      label = "Constant treatment effect"
+    ),
+    aes(
+      x     = x, 
+      color = label,
+      linetype = label
+    ),
+    size = .8,
+    alpha = .8,
+    fun  = predictBenefitModelBasedHte,
+    args = list(
+      modelBasedFit = constantModel
+    )
+  ) +
+  stat_function(
+    data = data.frame(
+      x = c(0, maxRisk), 
+      label = "Linear interaction"
+    ),
+    aes(
+      x     = x,
+      linetype = label,
+      color = label
+    ),
+    size = .65,
+    fun  = predictBenefitModelBasedHte,
+    args = list(
[…]

### 2022-04-12 code/InteractionPlots.R

@@ -1,0 +1,89 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+
+source("code/helpers/CreateManuscriptPlots.R")
+
+interactionIds <- c(10, 64, 66, 65)
+
+metric <- "rmse"
+titles <- c(
+  "Base case scenario",
+  "Weak interactions",
+  "Mixed interactions",
+  "Strong interactions"
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+f <- function(x) x * 100
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) %>% 
+  mutate_at(
+    c(
+      "constant_treatment_effect",
+      "stratified",
+      "linear_predictor",
+      "rcs_3_knots",
+      "rcs_4_knots",
+      "rcs_5_knots",
+      "adaptive"
+    ),
+    f
+  )
+
+plotList <- list()
+
+for (i in seq_along(interactionIds)) {
+  tmp <- processed %>%
+    dplyr::filter(scenarioId == interactionIds[i])
+  
+  plot <- createPlot(
+    data = tmp,
+    metric = metric,
+    title = titles[i],
+    limits = c(0, 10),
+    pointSize = .5
+  ) +
+    ggplot2::theme(
+      legend.position = "none",
+      axis.title.x    = ggplot2::element_blank(),
+      axis.text.x     = ggplot2::element_text(size = 9, angle = 45, hjust = 1),
+      axis.title.y    = ggplot2::element_blank(),
+      axis.text.y     = ggplot2::element_text(size = 9),
+      plot.title      = ggplot2::element_text(size = 12)
+    )
+  
+  plotList[[i]] <- plot
+}
+
+pp <- gridExtra::grid.arrange(
+  plotList[[1]],
+  plotList[[2]],
+  plotList[[3]],
+  plotList[[4]],
+  nrow = 2,
+  ncol = 2,
+  left = grid::textGrob(
+    expression(
+      paste(
+        "Root mean squared error (x ", 
+        10^-2, 
+        ")"
+      )
+    ),
+    rot = 90
+  )
+)
+  ggplot2::ggsave(
+    file.path("figures", "rmse_interactions.png"), 
+    plot = pp,
+    width = 8.5, 
+    height = 7
+  )

### 2022-04-12 code/MakeBook.R

@@ -1,0 +1,3 @@
+#!/usr/bin/env Rscript
+setwd("extras/bookdown")
+bookdown::render_book("index.Rmd", "bookdown::bs4_book", quiet = TRUE, output_dir = "../../docs")

### 2022-04-12 code/MergeResults.R

@@ -1,0 +1,16 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+source("code/helpers/TmpFiles.R")
+
+scenarios <- length(list.files("data/raw", pattern = "scenario"))
+if (!dir.exists(".scratch/tmp")) {
+  dir.create(".scratch/tmp", recursive = TRUE)
+}
+lapply(1:scenarios, createTmpFiles, scenarioDir = "data/raw/", tmpDir = ".scratch/tmp/")
+lapply(
+  c("rmse", "discrimination", "calibration", "concordance", "adaptiveModel"), 
+  mergeTmpFiles, 
+  tmpDir = ".scratch/tmp",
+  saveDir = "data/processed"
+  )

### 2022-04-12 code/NPatientsAucPlots.R

@@ -1,0 +1,90 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+library(ggtext)
+library(glue)
+
+source("code/helpers/CreateManuscriptPlots.R")
+
+interactionIds <- c(59, 60, 62, 63)
+
+metric <- "rmse"
+titles <- c(
+  "**A.** N: 1065, AUC: 0.65",
+  "**B.** N: 1065, AUC: 0.85",
+  "**C.** N: 17000, AUC: 0.65",
+  "**D.** N: 17000, AUC: 0.85"
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) 
+
+plotList <- list()
+
+for (i in seq_along(interactionIds)) {
+  tmp <- processed %>%
+    dplyr::filter(scenarioId == interactionIds[i])
+  
+  plot <- createPlot(
+    data = tmp,
+    metric = metric,
+    title = titles[i],
+    limits = c(0, .11),
+    pointSize = .5
+  ) 
+  
+  if (i %% 2 == 1)  {
+    plot <- plot +
+      ggplot2::theme(
+        legend.position = "none",
+        axis.title.x    = ggplot2::element_blank(),
+        axis.text.x     = ggplot2::element_text(size = 9, angle = 45, hjust = 1),
+        axis.title.y    = ggplot2::element_text(size = 11),
+        axis.text.y     = ggplot2::element_text(size = 9),
+        plot.title      = ggtext::element_markdown(size = 12)
+      )
+  } else {
+    plot <- plot +
+      ggplot2::theme(
+        legend.position = "none",
+        axis.title.x    = ggplot2::element_blank(),
+        axis.text.x     = ggplot2::element_text(size = 9, angle = 45, hjust = 1),
+        axis.title.y    = ggplot2::element_blank(),
+        axis.text.y     = ggplot2::element_text(size = 9),
+        plot.title      = ggtext::element_markdown(size = 12)
+      )
+  }
+  
+  plotList[[i]] <- plot
+}
+
+pp <- gridExtra::grid.arrange(
+  plotList[[1]],
+  plotList[[2]],
+  plotList[[3]],
+  plotList[[4]],
+  nrow = 2,
+  ncol = 2
+)
+  ggplot2::ggsave(
[…]

### 2022-04-12 code/NlAucPlots.R

@@ -1,0 +1,88 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+
+source("code/helpers/CreateManuscriptPlots.R")
+
+interactionIds <- c(47, 48, 56, 57)
+
+metric <- "rmse"
+titles <- c(
+  "**A.** Non-linearity: Moderate, AUC: 0.65",
+  "**B.** Non-linearity: Moderate, AUC: 0.85",
+  "**C.** Non-linearity: High, AUC: 0.65",
+  "**D.** Non-linearity: High, AUC: 0.85"
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) 
+
+plotList <- list()
+
+for (i in seq_along(interactionIds)) {
+  tmp <- processed %>%
+    dplyr::filter(scenarioId == interactionIds[i])
+  
+  plot <- createPlot(
+    data = tmp,
+    metric = metric,
+    title = titles[i],
+    limits = c(0, .11),
+    pointSize = .5
+  ) 
+  
+  if (i %% 2 == 1)  {
+    plot <- plot +
+      ggplot2::theme(
+        legend.position = "none",
+        axis.title.x    = ggplot2::element_blank(),
+        axis.text.x     = ggplot2::element_text(size = 9, angle = 45, hjust = 1),
+        axis.title.y    = ggplot2::element_text(size = 11),
+        axis.text.y     = ggplot2::element_text(size = 9),
+        plot.title      = ggtext::element_markdown(size = 12)
+      )
+  } else {
+    plot <- plot +
+      ggplot2::theme(
+        legend.position = "none",
+        axis.title.x    = ggplot2::element_blank(),
+        axis.text.x     = ggplot2::element_text(size = 9, angle = 45, hjust = 1),
+        axis.title.y    = ggplot2::element_blank(),
+        axis.text.y     = ggplot2::element_text(size = 9),
+        plot.title      = ggtext::element_markdown(size = 12)
+      )
+  }
+  
+  plotList[[i]] <- plot
+}
+
+pp <- gridExtra::grid.arrange(
+  plotList[[1]],
+  plotList[[2]],
+  plotList[[3]],
+  plotList[[4]],
+  nrow = 2,
+  ncol = 2
+)
[…]

### 2022-04-12 code/NonLinearPlots.R

@@ -1,0 +1,77 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+
+source("code/helpers/CreateManuscriptPlots.R")
+
+metric <- "rmse"
+titles <- c(
+  "Linear deviation",
+  "Quadratic deviation"
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) 
+
+nonConstantIds <- c(37, 55)
+
+for (i in seq_along(nonConstantIds)) {
+  tmp <- processed %>%
+    dplyr::filter(scenarioId == nonConstantIds[i])
+  
+  plotName <- paste(metric, nonConstantIds[i], sep = "_")
+  
+  plot <- createPlot(
+    data = tmp,
+    metric = metric,
+    title = titles[i],
+    limits = c(0, .08)
+  )
+  
+  if (i > 1) {
+    plot <- plot +
+      ggplot2::theme(
+        legend.position = "none",
+        axis.title.x    = ggplot2::element_blank(),
+        axis.title.y    = ggplot2::element_blank(),
+        axis.text.x     = ggplot2::element_text(size = 16, angle = 45, hjust = 1),
+        axis.text.y     = ggplot2::element_text(size = 16),
+        plot.title      = ggplot2::element_text(size = 22)
+      )
+  } else {
+    plot <- plot +
+      ggplot2::theme(
+        legend.position = "none",
+        axis.title.x    = ggplot2::element_blank(),
+        axis.title.y    = ggplot2::element_text(size = 18),
+        axis.text.x     = ggplot2::element_text(size = 16, angle = 45, hjust = 1),
+        axis.text.y     = ggplot2::element_text(size = 16),
+        plot.title      = ggplot2::element_text(size = 22)
+      )
+  }
+  
+  ggplot2::ggsave(
+    plot = plot,
+    file.path("figures", paste0(plotName, ".tiff")),
+    width  = 8.5,
+    height = 7,
+    device = "tiff",
+    compression = "lzw"
+  )
+  
+  ggplot2::ggsave(
+    plot = plot,
+    file.path("figures", paste0(plotName, ".png")),
+    width  = 8.5,
+    height = 7,
+    device = "png"
+  )
+  
+}

### 2022-04-12 code/NonLinearityNPatientsPlots.R

@@ -1,0 +1,88 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+
+source("code/helpers/CreateManuscriptPlots.R")
+
+interactionIds <- c(49, 52, 58, 61)
+
+metric <- "rmse"
+titles <- c(
+  "**A.** Non-linearity: Moderate, N: 1064",
+  "**B.** Non-linearity: Moderate, N: 17000",
+  "**C.** Non-linearity: High, N: 1064",
+  "**D.** Non-linearity: High, N: 17000"
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) 
+
+plotList <- list()
+
+for (i in seq_along(interactionIds)) {
+  tmp <- processed %>%
+    dplyr::filter(scenarioId == interactionIds[i])
+  
+  plot <- createPlot(
+    data = tmp,
+    metric = metric,
+    title = titles[i],
+    limits = c(0, .11),
+    pointSize = .5
+  ) 
+  
+  if (i %% 2 == 1)  {
+    plot <- plot +
+      ggplot2::theme(
+        legend.position = "none",
+        axis.title.x    = ggplot2::element_blank(),
+        axis.text.x     = ggplot2::element_text(size = 9, angle = 45, hjust = 1),
+        axis.title.y    = ggplot2::element_text(size = 11),
+        axis.text.y     = ggplot2::element_text(size = 9),
+        plot.title      = ggtext::element_markdown(size = 12)
+      )
+  } else {
+    plot <- plot +
+      ggplot2::theme(
+        legend.position = "none",
+        axis.title.x    = ggplot2::element_blank(),
+        axis.text.x     = ggplot2::element_text(size = 9, angle = 45, hjust = 1),
+        axis.title.y    = ggplot2::element_blank(),
+        axis.text.y     = ggplot2::element_text(size = 9),
+        plot.title      = ggtext::element_markdown(size = 12)
+      )
+  }
+  
+  plotList[[i]] <- plot
+}
+
+pp <- gridExtra::grid.arrange(
+  plotList[[1]],
+  plotList[[2]],
+  plotList[[3]],
+  plotList[[4]],
+  nrow = 2,
+  ncol = 2
+)
+  ggplot2::ggsave(
[…]

### 2022-04-12 code/PlotDeviations.R

@@ -1,0 +1,166 @@
+#!/usr/bin/env Rscript
+
+# Description:
+#   Plots the theoretical deviations from the base case scenario
+# Output:
+#   figures/deviate_linear_08.png
+#   figures/deviate_quadratic_08.png
+#   figures/deviate_linear_absolute_08.png
+#   figures/deviate_quadratic_absolute_08.png
+# Depends:
+#   code/PlotGammas.R
+
+
+
+source("code/helpers/PlotGammas.R")
+
+library(tidyverse)
+library(SmoothHte)
+library(SimulateHte)
+library(SimulationEvaluationHte)
+library(dplyr)
+
+analysisIds <- readr::read_csv(
+  "data/processed/analysisIds.csv"
+  )
+
+analysisIdsInteractions <- readr::read_csv(
+  "data/processed/analysisIdsInteractions.csv",
+  col_types = "icfiidddddddddddddd"
+)
+
+generatePlot <- function(scenarios, analysisIds) {
+  
+  gammas <- deviationPlots <- list()
+  idSettings <- analysisIds %>% filter(scenario %in% scenarios)
+  for (i in seq_along(scenarios)) {
+    gammas[[i]] <- list(
+      g0 = idSettings %>% filter(scenario == scenarios[i]) %>% pull(g0),
+      g1 = idSettings %>% filter(scenario == scenarios[i]) %>% pull(g1),
+      g2 = idSettings %>% filter(scenario == scenarios[i]) %>% pull(g2),
+      c  = idSettings %>% filter(scenario == scenarios[i]) %>% pull(c)
+    )
+  }
+
+  names(gammas) <- sub("^[lq].*-", "", idSettings$type)
+  
+  initialPlot <- constantEffectPlot(or = .8, label = "base-case") +
+    plotSquare(g0 = 0, g1 = 1, label = "no effect", linetype = "dashed")
+  types <- names(gammas)
+
+
+  deviationPlots <- map(
+    types,
+    function(type) {
+      plotSquare(
+        g0 = gammas[[type]]$g0,
+        g1 = gammas[[type]]$g1,
+        g2 = gammas[[type]]$g2,
+        l = gammas[[type]]$c,
+        label = type
+      )
+    }
+  )
+  p <- initialPlot +
+    deviationPlots[[1]]
+  if (length(types) > 1) {
+    p <- p + deviationPlots[[2]]
+  }
+  p <- p +
+    xlab("Linear predictor in control arm") +
[…]

### 2022-04-12 code/PlotDeviationsManuscript.R

@@ -1,0 +1,57 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+source("code/helpers/PlotGammas.R")
+source("code/helpers/PlotDeviationsFunctions.R")
+
+library(tidyverse)
+library(SmoothHte)
+library(SimulateHte)
+library(SimulationEvaluationHte)
+library(dplyr)
+
+analysisIds <- readr::read_csv(
+  "data/processed/analysisIds.csv",
+  col_types = "iffiiddddddddddddd"
+)
+
+analysisIdsInteractions <- readr::read_csv(
+  "data/processed/analysisIdsInteractions.csv",
+  col_types = "icfiidddddddddddddd"
+)
+
+scenarios               <- c(10, 28, 37, 55)
+plotsManuscript         <- generatePlot(scenarios = scenarios)
+plotsAbsoluteManuscript <- generateAbsolutePlot(scenarios)
+
+ggsave(
+  filename    = "figures/deviationsManuscript.png", 
+  plot        = plotsManuscript,
+  width       = 12,
+  height      = 12,
+  units       = "cm"
+)
+ggsave(
+  filename    = "figures/deviationsManuscript.tiff", 
+  plot        = plotsManuscript,
+  dpi         = 1200,
+  device      = "tiff",
+  compression = "lzw"
+)
+ggsave(
+  filename    = "figures/deviationsAbsoluteManuscript.png", 
+  plot        = plotsAbsoluteManuscript,
+  width       = 12,
+  height      = 12,
+  units       = "cm"
+)
+ggsave(
+  filename    = "figures/deviationsAbsoluteManuscript.tiff", 
+  plot        = plotsAbsoluteManuscript,
+  dpi         = 1200,
+  device      = "tiff",
+  compression = "lzw"
+)

### 2022-04-12 code/PlotRmse.R

@@ -1,0 +1,380 @@
+#!/usr/bin/env Rscript
+
+# ===================================================
+# Description:
+#   Generates the RMSE plots for the base case
+# Input:
+#   - sample size
+#   - auc
+#   - value
+# Output:
+#   - figures/rmse_base.tiff
+# Depends:
+#   - data/processed/analysisIds.csv
+#   - data/processed/rmse.csv
+#   - code/helpers/CreateManuscriptPlots.R
+#   - code/helpers/PlotResult.R
+#   - code/helpers/Absolute.R
+# ===================================================
+
+args <- commandArgs(trailingOnly = TRUE)
+args_base <- as.character(args[1])
+args_sampleSize <- as.numeric(args[2])
+args_auc <- as.numeric(args[3])
+args_value <- as.character(args[4])
+
+library(tidyverse)
+library(glue)
+library(ggtext)
+library(gridExtra)
+library(grid)
+library(ggside)
+
+# --------------------------------------
+# Sourcing helper files:
+#   1. Generates single boxplot
+#   2. Generates the boxplot list
+#   3. Generates the absolute plots
+# --------------------------------------
+source("code/helpers/CreateManuscriptPlots.R") 
+source("code/helpers/PlotResult.R")
+source("code/helpers/Absolute.R")
+
+scenarioIds <- readr::read_csv("data/processed/analysisIds.csv") %>%
+  filter(
+    base == args_base,
+    !(type %in% c("quadratic-moderate", "linear-moderate")),
+    sampleSize == args_sampleSize,
+    auc == args_auc
+  ) 
+
+metric    <- "rmse"
+
+titles <- scenarioIds %>%
+  mutate(
+    title = ifelse(
+      type == "constant",
+      str_to_sentence(glue("{str_replace_all(type, '-', ' ')} treatment effect")),
+      str_to_sentence(glue("{str_replace_all(type, '-', ' ')} deviation"))
+    )
+  ) %>%
+  select(title) %>%
+  unlist() %>%
+  unique()
+
+names(titles) <- NULL
+
+titlePrefix <- paste0(
+  "**",
+  LETTERS[1:5],
+  ".**"
+)
+
+titles <- paste(
+  titlePrefix,
+  titles
+)
+
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+f <- function(x) x * 100
+
+limitsHigh <- ifelse(
+  args_base == "high",
+  yes = .2,
+  no = .15
+)
+
+processed <- readr::read_csv(
[…]

### 2022-04-12 code/PlotRmseAbsent.R

@@ -1,0 +1,100 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+source("code/helpers/CreateManuscriptPlots.R")
+source("code/helpers/PlotResult.R")
+
+scenarioIds <- readr::read_csv("data/processed/analysisIds.csv")
+
+scenarios <- c(
+  5, 4, 6,
+  2, 1, 3,
+  8, 7, 9
+)
+metric    <- "rmse"
+value     <- "absent"
+
+titles <- scenarioIds %>%
+  filter(scenario %in% scenarios) %>%
+  select(title) %>%
+  unlist()
+
+names(titles) <- NULL
+
+titlePrefix <- paste0(
+  "**",
+  LETTERS[1:4],
+  ".**"
+)
+
+titles <- paste(
+  titlePrefix,
+  titles
+)
+
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+f <- function(x) x * 100
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) %>% 
+  mutate_at(
+    c(
+      "constant_treatment_effect",
+      "stratified",
+      "linear_predictor",
+      "rcs_3_knots",
+      "rcs_4_knots",
+      "rcs_5_knots",
+      "adaptive"
+    ),
+    f
+  )
+
+plotList <- plotResult(scenarios, processed, titles = NULL)
+
+pp <- gridExtra::grid.arrange(
+  plotList[[1]],
+  plotList[[2]],
+  plotList[[3]],
+  plotList[[4]],
+  plotList[[5]],
+  plotList[[6]],
+  plotList[[7]],
+  plotList[[8]],
+  plotList[[9]],
+  nrow = 3,
+  ncol = 3,
+  left = grid::textGrob(
+    "Sample size",
+    rot = 90
+  ),
+  top = grid::textGrob(
+    "Prediction AUC"
+  )
+)
+
+fileName <- paste0(
+  paste(
+    metric,
+    value,
+    sep = "_"
+  ),
+  ".tiff"
+)
+  ggplot2::ggsave(
+    file.path("figures", fileName), 
+    plot = pp,
+    dpi = 1200,
+    width = 10.5, 
+    height = 8,
+    compression = "lzw"
+  )
+  
+  

### 2022-04-12 code/PlotRmseInteractions.R

@@ -1,0 +1,439 @@
+#!/usr/bin/env Rscript
+
+# ===================================================
+# Description:
+#   Generates the RMSE plots for the base case
+# Input:
+#   - type
+#   - value
+# Output:
+#   - figures/rmse_base.tiff
+# Depends:
+#   - data/processed/analysisIds.csv
+#   - data/processed/rmse.csv
+#   - code/helpers/CreateManuscriptPlots.R
+#   - code/helpers/PlotResult.R
+#   - code/helpers/Absolute.R
+# ===================================================
+
+library(tidyverse)
+library(glue)
+library(ggtext)
+library(gridExtra)
+library(grid)
+library(ggside)
+
+args <- commandArgs(trailingOnly = TRUE)
+interactionType <- as.character(args[1])
+
+if (interactionType == "positive") {
+  scenarioLimits <- c(649, 660)
+} else if (interactionType == "negative") {
+  scenarioLimits <- c(661, 672)
+} else {
+  scenarioLimits <- c(673, 676)
+}
+
+
+# --------------------------------------
+# Sourcing helper files:
+#   1. Generates single boxplot
+#   2. Generates the boxplot list
+#   3. Generates the absolute plots
+# --------------------------------------
+source("code/helpers/CreateManuscriptPlots.R")     
+source("code/helpers/PlotResult.R")                
+source("code/helpers/Absolute.R")                  
+
+scenarioIds <- readr::read_csv("data/processed/analysisIdsInteractions.csv") %>%
+  filter(scenario >= scenarioLimits[1] & scenario <= scenarioLimits[2])
+
+metric    <- "rmse"
+titles <- scenarioIds %>%
+  mutate(
+    title = str_to_sentence(glue("{str_replace_all(type, '-', ' ')} interactions")
+    )
+  ) %>%
+  select(title) %>%
+  unlist() %>%
+  unique()
+
+if (interactionType != "combined") {
+  titlePrefix <- paste0(
+    "**",
+    LETTERS[1:3],
+    ".**"
+  )
+} else {
+  titlePrefix <- ""
+}
+
+titles <- paste(
+  titlePrefix,
+  titles
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+f <- function(x) x * 100
+
+limitsHigh <- .15
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) %>% 
[…]

### 2022-04-12 code/PlotScenarioQuartileBoxplots.R

@@ -1,0 +1,142 @@
+#!/usr/bin/env Rscript
+
+# ==========================================================================
+# Description:
+#   Generates boxplots of simulation scenarios that resemble the plots
+#   shown in Kent, et al. 2016
+# Input:
+#   - scenario id
+# Output:
+#   - figures/scenario_XXX.tiff
+# Depends:
+#   - data/processed/analysisIds.csv
+# ==========================================================================
+
+set.seed(19910930)
+
+library(tidyverse)
+library(SimulateHte)
+
+args <- commandArgs(trailingOnly = TRUE)
+selectedScenario <- as.integer(args[1])
+
+analysisIds <- readr::read_csv("data/processed/analysisIds.csv")
+
+idSettings <- analysisIds %>%
+  filter(scenario == selectedScenario)
+
+if (idSettings$base != "absent") {
+  harm <- case_when(
+    idSettings$harm == "moderate-positive" ~ idSettings$averageTrueBenefit / 4, 
+    idSettings$harm == "strong-positive"   ~ idSettings$averageTrueBenefit / 2, 
+    idSettings$harm == "negative"          ~ -idSettings$averageTrueBenefit / 4, 
+    TRUE                                   ~ 0
+  )
+} else {
+  harm <- case_when(
+    idSettings$harm == "moderate-positive" ~ .01,
+    idSettings$harm == "strong-positive"   ~ .02,
+    idSettings$harm == "negative"          ~ -.01,
+    TRUE                                   ~ 0
+  )
+}
+  
+  createF1 <- function(c) function(x) x - c
+  createF2 <- function(c) function(x) (x - c)^2
+  treatmentEffectSettings <- SimulateHte::createTreatmentEffectSettings(
+    type = "lp",
+    harm = harm,
+    modelSettings = SimulateHte::createModelSettings(
+      constant = idSettings$g0,
+      modelMatrix = matrix(c(1, 1)),
+      transformationSettings = list(
+        createF1(idSettings$c),
+        createF2(idSettings$c)
+      ),
+      coefficients = c(
+        idSettings$g1,
+        idSettings$g2
+      )
+    )
+  )
+databaseSettings <- SimulateHte::createDatabaseSettings(
+  numberOfObservations = 4e4,
+  numberOfCovariates = 8,
+  covariateDistributionSettings = list(
[…]

### 2022-04-12 code/RmseConstant.R

@@ -1,0 +1,97 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+
+source("code/helpers/CreateManuscriptPlots.R")
+
+scenarios <- c(10, 12, 16)
+
+metric <- "rmse"
+titles <- c(
+  "**A.** Constant treatment effect",
+  "**B.** AUC: 0.85; N: 4250",
+  "**C.** AUC: 0.75; N: 17000"
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+f <- function(x) x * 100
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) %>% 
+  mutate_at(
+    c(
+      "constant_treatment_effect",
+      "stratified",
+      "linear_predictor",
+      "rcs_3_knots",
+      "rcs_4_knots",
+      "rcs_5_knots",
+      "adaptive"
+    ),
+    f
+  )
+
+plotList <- list()
+
+for (i in seq_along(scenarios)) {
+  tmp <- processed %>%
+    dplyr::filter(scenarioId == scenarios[i])
+  
+  plot <- createPlot(
+    data = tmp,
+    metric = metric,
+    title = titles[i],
+    limits = c(0, 6),
+    pointSize = .5
+  ) +
+    ggplot2::theme(
+      legend.position = "none",
+      axis.title.x    = ggplot2::element_blank(),
+      axis.text.x     = ggplot2::element_text(size = 9, angle = 45, hjust = 1),
+      axis.title.y    = ggplot2::element_blank(),
+      axis.text.y     = ggplot2::element_text(size = 9),
+      plot.title      = ggtext::element_markdown(size = 12)
+    )
+  
+  plotList[[i]] <- plot
+}
+
+pp <- gridExtra::grid.arrange(
+  plotList[[1]],
+  plotList[[2]],
+  plotList[[3]],
+  nrow = 1,
+  ncol = 3,
+  left = grid::textGrob(
+    expression(
+      paste(
+        "Root mean squared error (x ", 
+        10^-2, 
+        ")"
+      )
+    ),
+    rot = 90
+  )
+)
+  ggplot2::ggsave(
+    file.path("figures", "rmse_constant.tiff"), 
+    plot = pp,
+    dpi = 1200,
+    width = 11, 
+    height = 4,
+    compression = "lzw"
+  )
+  
[…]

### 2022-04-12 code/SampleSizePlots.R

@@ -1,0 +1,78 @@
+#!/usr/bin/env Rscript
+
+# Description: 
+# Output: 
+# Depends: 
+
+library(tidyverse)
+
+source("code/helpers/CreateManuscriptPlots.R")
+
+metric <- "rmse"
+titles <- c(
+  "N = 1,064",
+  "N = 17,000"
+)
+
+metricFile <- paste(metric, "csv", sep = ".")
+
+processed <- readr::read_csv(
+  file = file.path("data/processed", metricFile)
+) 
+
+nonConstantIds <- c(58, 61)
+
+for (i in seq_along(nonConstantIds)) {
+  tmp <- processed %>%
+    dplyr::filter(scenarioId == nonConstantIds[i])
+  
+  plotName <- paste(metric, nonConstantIds[i], sep = "_")
+  
+  plot <- createPlot(
+    data = tmp,
+    metric = metric,
+    title = titles[i],
+    limits = c(0, .1)
+  )
+  
+  if (i > 1) {
+    plot <- plot +
+      ggplot2::theme(
+        legend.position = "none",
+        axis.title.x    = ggplot2::element_blank(),
+        axis.title.y    = ggplot2::element_blank(),
+        axis.text.x     = ggplot2::element_text(size = 16, angle = 45, hjust = 1),
+        axis.text.y     = ggplot2::element_text(size = 16),
+        plot.title      = ggplot2::element_text(size = 22)
+      )
+  } else {
+    plot <- plot +
+      ggplot2::theme(
+        legend.position = "none",
+        axis.title.x    = ggplot2::element_blank(),
+        axis.title.y    = ggplot2::element_text(size = 18),
+        axis.text.x     = ggplot2::element_text(size = 16, angle = 45, hjust = 1),
+        axis.text.y     = ggplot2::element_text(size = 16),
+        plot.title      = ggplot2::element_text(size = 22)
+      )
+  }
+  
+    ggplot2::ggsave(
+      filename = paste0(plotName, ".png"),
+      plot   = plot,
+      path   = "figures",
+      width  = 8.5,
+      height = 7
+    )
+    
+    ggplot2::ggsave(
+      filename    = paste0(plotName, ".tiff"),
+      plot        = plot,
+      path        = "figures",
+      width       = 8.5,
+      height      = 7,
+      device      = "tiff",
+      compression = "lzw"
+    )
+  
+}

### 2022-04-12 code/SelectedModelAdaptive.R

@@ -1,0 +1,134 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(ggthemes)
+
+args <- commandArgs(trailingOnly = TRUE)
+args_base <- as.character(args[1])
+args_sampleSize <- as.numeric(args[2])
+args_auc <- as.numeric(args[3])
+args_value <- as.character(args[4])
+
+analysisIds <- read_csv("data/processed/analysisIds.csv")
+
+scenarios <- analysisIds %>%
+  filter(
+    base == args_base,
+    !(type %in% c("quadratic-moderate", "linear-moderate")),
+    harm != "negative",
+    sampleSize == args_sampleSize,
+    auc == args_auc
+  ) %>%
+  pull(scenario)
+
+adaptiveModel <- readr::read_csv("data/processed/adaptiveModel.csv") %>%
+  filter(scenarioId %in% scenarios) %>%
+  left_join(analysisIds, by = c("scenarioId" = "scenario")) %>%
+  select(
+    c("selectedAdaptiveModel", "type", "harm")
+  ) %>%
+  mutate(
+    selectedAdaptiveModel = factor(
+      selectedAdaptiveModel,
+      levels = rev(
+        c(
+          "treatment",
+          "risk",
+          "RCS with 3 knots",
+          "RCS with 4 knots",
+          "RCS with 5 knots"
+        )
+      ),
+      labels = rev(
+        c(
+          "Constant treatment effect",
+          "Linear interaction",
+          "RCS-3",
+          "RCS-4",
+          "RCS-5"
+        )
+      )
+    ),
+    type = factor(
+      type,
+      levels = c(
+        "constant",
+        "linear-high",
+        "quadratic-high",
+        "non-monotonic"
+      ),
+      labels = c(
+        "Constant treatment effect",
+        "Strong linear deviation",
+        "Strong quadratic deviation",
+        "Non-monotonic deviation"
+      )
+    )
+  )
+
+p <- ggplot2::ggplot(
+  data = adaptiveModel,
+  ggplot2::aes(x = harm, fill = selectedAdaptiveModel)
+) +
+  ggplot2::geom_bar(position = "fill") +
+  scale_fill_manual(
+    values = c(
+      "#264653",
+      "#2A9D8F",
+      "#E9C46A",
+      "#F4A261",
+      "#E76F51"
+    ),
+    breaks = c(
+      "Constant treatment effect",
+      "Linear interaction",
[…]

### 2022-04-12 code/SimulationScript.R

@@ -1,0 +1,232 @@
+#!/usr/bin/env Rscript
+
+# Description:
+#   Script to be run for each simulation scenario
+# Depends:
+#   data/processed/analysisIds.rds
+# Output:
+#   data/processed/scenario_#/
+# Note:
+#   Needs a data/raw/scenario_# as input
+
+args <- commandArgs(trailingOnly = TRUE)
+if (length(args) != 1) stop("Requires scenario id")
+
+library(tidyverse)
+library(SmoothHte)
+library(SimulateHte)
+library(SimulationEvaluationHte)
+library(dplyr)
+
+analysisIds <- readr::read_csv(
+  "data/processed/analysisIds.csv",
+  col_types =  cols(
+    .default = col_double(),
+    base = col_character(),
+    type = col_character(),
+    harm = col_character()
+  )
+)
+
+analysisIdsInteractions <- readr::read_csv(
+  "data/processed/analysisIdsInteractions.csv",
+)
+
+selectedScenario <- as.numeric(
+  stringr::str_extract(args[1], "[0-9]+")
+)
+
+if (selectedScenario <= 648) {
+  idSettings <- analysisIds %>%
+    filter(scenario == selectedScenario)
+  
+  if (idSettings$base != "absent") {
+    harm <- case_when(
+      idSettings$harm == "moderate-positive" ~ idSettings$averageTrueBenefit / 4, 
+      idSettings$harm == "strong-positive"   ~ idSettings$averageTrueBenefit / 2, 
+      idSettings$harm == "negative"          ~ -idSettings$averageTrueBenefit / 4, 
+      TRUE                                   ~ 0
+    )
+  } else {
+    harm <- case_when(
+      idSettings$harm == "moderate-positive" ~ .01,
+      idSettings$harm == "strong-positive"   ~ .02,
+      idSettings$harm == "negative"          ~ -.01,
+      TRUE                                   ~ 0
+    )
+  }
+  
+  createF1 <- function(c) function(x) x - c
+  createF2 <- function(c) function(x) (x - c)^2
+  treatmentEffectSettings <- SimulateHte::createTreatmentEffectSettings(
+    type = "lp",
+    harm = harm,
+    modelSettings = SimulateHte::createModelSettings(
+      constant = idSettings$g0,
+      modelMatrix = matrix(c(1, 1)),
+      transformationSettings = list(
+        createF1(idSettings$c),
+        createF2(idSettings$c)
+      ),
+      coefficients = c(
[…]

### 2022-04-12 code/WriteAnalysisIds.R

@@ -1,0 +1,164 @@
+#!/usr/bin/env Rscript
+
+# Description: Generates the analysisIds
+# Output: data/processed/analysisIds.csv
+# Depends: 
+
+library(tidyverse)
+library(readr)
+
+base <- c(
+  "absent",
+  "moderate",
+  "high"
+)
+
+type <- c(
+  "constant",
+  "linear-moderate",
+  "linear-high",
+  "quadratic-moderate",
+  "quadratic-high",
+  "non-monotonic"
+)
+
+auc <- c(.75, .65, .85)
+
+sampleSize <- c(4250, 1063, 17000)
+harm <- c("absent", "moderate-positive", "strong-positive", "negative")
+
+expand_grid(base, type, sampleSize, auc, harm) %>%
+  mutate(
+    b0 = case_when(
+      auc == .75 ~ -2.08,
+      auc == .65 ~ -1.63,
+      auc == .85 ~ -2.7,
+      TRUE ~ 0
+    ),
+    b1 = case_when(
+      auc == .75 ~ .49,
+      auc == .65 ~ .26,
+      auc == .85 ~ .82,
+      TRUE ~ 0
+    ),
+    b2 = b1,
+    b3 = b1,
+    b4 = b1,
+    b5 = b1,
+    b6 = b1,
+    b7 = b1,
+    b8 = b1,
+    g0 = case_when(
+      base == "absent" & type == "constant" ~ 0,
+      base == "moderate" & type == "constant" ~ log(.8),
+      base == "high" & type == "constant" ~ log(.5),
+      base == "absent" & type == "linear-moderate" & auc == .75 ~ -.06,
+      base == "absent" & type == "linear-moderate" & auc == .65 ~ -.08,
+      base == "absent" & type == "linear-moderate" & auc == .85 ~ -.07,
+      base == "absent" & type == "linear-high" & auc == .75 ~ -.25,
+      base == "absent" & type == "linear-high" & auc == .65 ~ -.29,
+      base == "absent" & type == "linear-high" & auc == .85 ~ -.22,
+      base == "absent" & type == "quadratic-moderate" & auc == .75 ~ -4.81,
+      base == "absent" & type == "quadratic-moderate" & auc == .65 ~ -4.77,
+      base == "absent" & type == "quadratic-moderate" & auc == .85 ~ -4.71,
+      base == "absent" & type == "quadratic-high" & auc == .75 ~ -4.22,
+      base == "absent" & type == "quadratic-high" & auc == .65 ~ -4.16,
[…]

### 2022-04-12 code/WriteAnalysisIdsInteractions.R

@@ -1,0 +1,84 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(readr)
+
+base <- "interaction"
+type <- c(
+  "weak", 
+  "mixed", 
+  "strong",
+  "negative-weak", 
+  "negative-mixed", 
+  "negative-strong",
+  "combined"
+)
+sampleSize <- 4250
+auc <- .75
+harm <- c("absent", "moderate-positive", "strong-positive", "negative")
+
+table <- expand_grid(base, type, sampleSize, auc, harm) %>%
+  mutate(
+    b0 = -2.08,
+    # b0 = case_when(
+    #   type == "weak" ~ -3.05,
+    #   type == "mixed" ~ -3.32,
+    #   TRUE ~ -3.50
+    # ),
+    b1 = .49,
+    b2 = .49,
+    b3 = .49,
+    b4 = .49,
+    b5 = .49,
+    b6 = .49,
+    b7 = .49,
+    b8 = .49,
+    g0 = log(.5),
+    g1 = case_when(
+      (type == "weak" | type == "mixed"                   ) ~ -.19,
+      (type == "strong" | type == "combined"              ) ~ -.49,
+      (type == "negative-weak" | type == "negative-mixed" ) ~ .16,
+      (type == "negative-strong"                          ) ~ .33
+    ),
+    g2 = case_when(
+      (type == "weak"                                                           ) ~ -.19,
+      (type == "mixed" | type == "strong"                                       ) ~ -.49,
+      (type == "negative-weak"                                                  ) ~ .16,
+      (type == "negative-mixed" | type == "negative-strong" | type == "combined") ~ .33
+    ),
+    g5 = case_when(
+      (type == "weak" | type == "mixed"                   ) ~ -.19,
+      (type == "strong" | type =="combined"               ) ~ -.49,
+      (type == "negative-weak" | type == "negative-mixed" ) ~ .16,
+      (type == "negative-strong"                          ) ~ .33
+    ),
+    g6 = case_when(
+      (type == "weak"                                                           ) ~ -.19,
+      (type == "mixed" | type == "strong"                                       ) ~ -.49,
+      (type == "negative-weak"                                                  ) ~ .16,
+      (type == "negative-mixed" | type == "negative-strong" | type == "combined") ~ .33
+    ),
+    averageTrueBenefit = case_when(
[…]

### 2022-04-12 code/helpers/Absolute.R

@@ -1,0 +1,126 @@
+plotAbsoluteBenefit <- function(data, projectDir, type) {
+  calcSquare <- function(x, g0 = 0, g1 = 0, g2 = 0, l = 0) {
+    ret <- g0 + g1 * (x - l) + g2 * (x - l)**2
+    return(ret)
+  }
+  calcAbsoluteBenefit <- function(p, g0 = 0, g1 = 0, g2 = 0, l = 0, harm) {
+    x <- log(p / (1 - p))
+    sq <- calcSquare(x, g0, g1, g2, l)
+    benefit <- plogis(x) - plogis(sq) - harm
+    return(benefit)
+  }
+  base <- data$base[1]
+  harmSettings <- unique(data$harm)
+  res <- ggplot()
+  validationDataset <- list()
+  for (i in seq_along(harmSettings)) {
+    scenarioSettings <- data %>%
+      dplyr::filter(harm == harmSettings[i]) %>%
+      dplyr::mutate(
+        harmValue = case_when(
+          harm == "moderate-positive" ~ .data$averageTrueBenefit[1] / 4, 
+          harm == "strong-positive"   ~ .data$averageTrueBenefit[1] / 2, 
+          harm == "negative"          ~ -.data$averageTrueBenefit[1] / 4, 
+          TRUE                        ~ 0
+        )
+      )
+    scenarioDir <- paste(
+      "scenario",
+      as.character(scenarioSettings$scenario),
+      sep = "_"
+    )
+    args <- list(
+      g0   = scenarioSettings$g0,
+      g1   = scenarioSettings$g1,
+      g2   = scenarioSettings$g2,
+      l    = scenarioSettings$c,
+      harm = scenarioSettings$harmValue
+    )
+    label <- harmSettings[i]
+    if (base != "interaction") {
+      res <- res +
+        ggplot2::stat_function(
+          data = data.frame(x = c(.05, .95), label = label),
+          ggplot2::aes(x = x, color = label, group = label),
+          fun = calcAbsoluteBenefit,
+          args = args
+        )
+    }
+    if (!missing(projectDir)) {
+      settings <- readRDS(
+        file.path(
+          projectDir,
+          "data",
+          "raw",
+          scenarioDir,
+          "settings.rds"
+        )
+      )
+      validationDatabaseSettings <- settings$simulationSettings$databaseSettings
+      validationDatabaseSettings$numberOfObservations <- 1e5
[…]

### 2022-04-12 code/helpers/CreateManuscriptPlots.R

@@ -1,0 +1,93 @@
+createPlot <- function(
+  data,
+  metric,
+  title,
+  limits,
+  pointSize = 1
+) {
+  
+  yAxis <- dplyr::case_when(
+    metric == "rmse"           ~ "RMSE",
+    metric == "discrimination" ~ "Discrimination",
+    metric == "calibration"    ~ "Calibration"
+  )
+  
+  tmp <- data %>%
+    mutate(
+      harm = factor(
+        harm,
+        levels = c(unique(.data$harm, ordered = TRUE))
+      )
+    ) %>%
+    dplyr::select(
+      c(
+        "harm",
+        "constant_treatment_effect", 
+        "stratified", 
+        "linear_predictor", 
+        "rcs_3_knots",
+        "rcs_4_knots",
+        # "rcs_5_knots", 
+        "adaptive"
+      )
+    ) %>% 
+    reshape2::melt()
+  levels(tmp$variable) <- case_when(
+    levels(tmp$variable) == "linear_predictor"          ~ "Linear",
+    levels(tmp$variable) == "constant_treatment_effect" ~ "Constant",
+    levels(tmp$variable) == "stratified"                ~ "Stratified",
+    levels(tmp$variable) == "rcs_3_knots"               ~ "RCS-3",
+    levels(tmp$variable) == "rcs_4_knots"               ~ "RCS-4",
+    # levels(tmp$variable) == "rcs_5_knots"               ~ "RCS-5",
+    levels(tmp$variable) == "adaptive"                  ~ "Adaptive",
+    TRUE                                                ~ levels(tmp$variable)
+  )
+  
+  
+  plot <- ggplot2::ggplot(
+    data = tmp,
+    ggplot2::aes(
+      x    = variable, 
+      y    = value, 
+      fill = harm
+    )
+  ) +
+    ggplot2::geom_boxplot(
+      outlier.size = pointSize,
+      width = .8,
+      lwd = .4,
+      fatten = .9
+    ) +
+    ggplot2::scale_fill_manual(
+      name = "Constant treatment-\n related harm",
+      values = c(
+        "#26547C",
+        "#06D6A0",
+        "#EF476F"
+      ),
+      breaks = c(
+        "absent",
+        "moderate_positive",
+        "strong_positive"
+      ),
+      labels = c(
+        "absent",
+        "moderate",
+        "strong"
+      )
+    ) +
+    ggplot2::ylab(yAxis) +
[…]

### 2022-04-12 code/helpers/GetMedians.R

@@ -1,0 +1,32 @@
+getMedian <- function(data) {
+  data %>%
+  summarise(
+    median_constant = median(constant_treatment_effect),
+    median_stratified = median(stratified),
+    median_linear = median(linear_predictor),
+    median_rcs_3 = median(rcs_3_knots),
+    median_rcs_4 = median(rcs_4_knots),
+  )
+}
+
+getLower <- function(data) {
+  data %>%
+  summarise(
+    lower_constant   = quantile(constant_treatment_effect, probs = .025),
+    lower_stratified = quantile(stratified, probs = .025),
+    lower_linear     = quantile(linear_predictor, probs = .025),
+    lower_rcs_3      = quantile(rcs_3_knots, probs = .025),
+    lower_rcs_4      = quantile(rcs_4_knots, probs = .025),
+  )
+}
+
+getUpper <- function(data) {
+  data %>%
+  summarise(
+    upper_constant   = quantile(constant_treatment_effect, probs = .975),
+    upper_stratified = quantile(stratified, probs = .975),
+    upper_linear     = quantile(linear_predictor, probs = .975),
+    upper_rcs_3      = quantile(rcs_3_knots, probs = .975),
+    upper_rcs_4      = quantile(rcs_4_knots, probs = .975),
+  )
+}

### 2022-04-12 code/helpers/PlotDeviationsFunctions.R

@@ -1,0 +1,154 @@
+generatePlot <- function(scenarios) {
+  
+  gammas <- list()
+  for (i in seq_along(scenarios)) {
+    idSettings <- analysisIds %>%
+      filter(scenario == scenarios[i])
+    gammas[[i]] <- list(
+      g0 = idSettings$g0,
+      g1 = idSettings$g1,
+      g2 = idSettings$g2,
+      c  = idSettings$c
+    )
+  }
+  
+  names(gammas) <- c(
+    "gammas_base_case",
+    "gammas_linear_moderate",
+    "gammas_linear_high",
+    "gammas_quadratic_high"
+  )
+  
+  p <- constantEffectPlot(or = .8, label = "base-case") +
+    plotSquare(g0 = 0, g1 = 1, label = "no effect", linetype = "dashed") +
+    plotSquare(
+      g0    = gammas$gammas_linear_moderate$g0,
+      g1    = gammas$gammas_linear_moderate$g1,
+      g2    = gammas$gammas_linear_moderate$g2,
+      l     = gammas$gammas_linear_moderate$c,
+      label = "linear-moderate"
+    ) +
+    plotSquare(
+      g0    = gammas$gammas_linear_high$g0,
+      g1    = gammas$gammas_linear_high$g1,
+      g2    = gammas$gammas_linear_high$g2,
+      l     = gammas$gammas_linear_high$c,
+      label = "linear-strong"
+    ) +
+    plotSquare(
+      g0    = gammas$gammas_quadratic_high$g0,
+      g1    = gammas$gammas_quadratic_high$g1,
+      g2    = gammas$gammas_quadratic_high$g2,
+      l     = gammas$gammas_quadratic_high$c,
+      label = "quadratic-strong"
+    ) +
+    xlab("Linear predictor in control arm") +
+    ylab("Linear predictor in treatment arm") +
+    xlim(-5, 2) +
+    scale_color_manual(
+      values = c(
+        "black",
+        "#66c2a5",
+        "#fc8d62",
+        "#8da0cb",
+        "#e78ac3"
+      ),
+      breaks = c(
+        "no effect",
+        "base-case",
+        "linear-moderate",
+        "linear-strong",
+        "quadratic-strong"
+      )
+    ) +
+    theme_classic() +
+    theme(
+      axis.text       = element_text(size = 10),
+      axis.title      = element_text(size = 11),
[…]

### 2022-04-12 code/helpers/PlotGammas.R

@@ -1,0 +1,85 @@
+calcSquare <- function(x, g0 = 0, g1 = 0, g2 = 0, l = 0) {
+  ret <- g0 + g1 * (x - l) + g2 * (x - l)**2
+  return(ret)
+}
+
+plotSquare <- function(
+  g0       = 0, 
+  g1       = 0, 
+  g2       = 0,
+  l        = 0,
+  linetype = 1,
+  label    = "Diverge"
+) {
+  ggplot2::stat_function(
+    data = data.frame(x = c(-5, 2)),
+    ggplot2::aes(x = x, color = label),
+    fun = calcSquare,
+    args = list(g0 = g0, g1 = g1, g2 = g2, l = l),
+    linetype = linetype
+  )
+}
+
+calcAbsoluteBenefit <- function(p, g0 = 0, g1 = 0, g2 = 0, l = 0) {
+  x <- log(p / (1 - p))
+  sq <- calcSquare(x, g0, g1, g2, l)
+  benefit <- plogis(x) - plogis(sq)
+  return(benefit)
+}
+
+plotAbsoluteBenefit <- function(g0 = 0, g1 = 1, g2 = 0, l = 0, label) {
+  ggplot2::stat_function(
+    data = data.frame(x = c(.05, .95)),
+    ggplot2::aes(x = x, color = label),
+    fun = calcAbsoluteBenefit,
+    args = list(g0 = g0, g1 = g1, g2 = g2, l = l)
+  )
+  
+}
+
+constantEffectPlot <- function(
+  or,
+  label = "constant"
+) {
+  p <- ggplot2::ggplot() +
+    ggplot2::stat_function(
+      data = data.frame(x = c(-5, 2)),
+      ggplot2::aes(x = x, color = label),
+      fun = calcSquare,
+      args = list(g0 = log(or), g1 = 1)
+    )
+  return(p)
+}
+
+# # plot linear -----------
+# gammas_moderate <- derive_gammas_linear(-.2, .04, -5, -.45, .8)
+# # gammas_high     <- derive_gammas_linear(-.4, .06, -5, -.45, .8)
+# gammas_high     <- derive_gammas_linear(-.8, .15, -5, -.45, .8)
+# constantEffectPlot(or = .8, label = "base-case") +
+#   plotSquare(g0 = 0, g1 = 1, label = "no effect", linetype = "dashed") +
+#   plotSquare(
+#     g0 = gammas_moderate$gamma_0,
+#     g1 = gammas_moderate$gamma_1,
[…]

### 2022-04-12 code/helpers/PlotResult.R

@@ -1,0 +1,28 @@
+plotResult <- function(scenarios, processed, titles, metric, limits = c(0, 10)) {
+  
+  plotList <- list()
+  for (i in seq_along(scenarios)) {
+    tmpList <- list()
+    tmpList$absent <- processed %>%
+      dplyr::filter(scenarioId == scenarios[i])
+    tmpList$moderate_positive <- processed %>%
+      dplyr::filter(scenarioId == scenarios[i] + 1)
+    tmpList$strong_positive <- processed %>%
+      dplyr::filter(scenarioId == scenarios[i] + 2)
+    
+    tmpData <- bind_rows(tmpList, .id = "harm")
+    
+    plot <- createPlot(
+      data = tmpData,
+      metric = metric,
+      title = titles[i],
+      limits = limits,
+      pointSize = .2
+    )
+    
+    plotList[[i]] <- plot
+  }
+  
+  return(plotList)
+}
+

### 2022-04-12 code/helpers/TmpFiles.R

@@ -1,0 +1,145 @@
+library(tidyverse)
+
+createTmpFiles <- function(scenarioId, scenarioDir, tmpDir) {
+  
+  evaluation <- readRDS(
+    file.path(
+      scenarioDir,
+      paste(
+        "scenario",
+        scenarioId,
+        sep = "_"
+      ),
+      "evaluation.rds"
+    )
+  )
+  
+  evaluation <- evaluation %>% map(~mutate(., scenarioId = scenarioId))
+  
+  readr::write_csv(
+    evaluation$rmse,
+    file.path(
+      tmpDir,
+      paste(
+        paste(
+          "tmp",
+          "rmse",
+          "scenario",
+          scenarioId,
+          sep = "_"
+        ),
+        "csv",
+        sep = "."
+      )
+    )
+  )
+  
+  readr::write_csv(
+    evaluation$discrimination,
+    file.path(
+      tmpDir,
+      paste(
+        paste(
+          "tmp",
+          "discrimination",
+          "scenario",
+          scenarioId,
+          sep = "_"
+        ),
+        "csv",
+        sep = "."
+      )
+    )
+  )
+  
+  readr::write_csv(
+    evaluation$calibration,
+    file.path(
+      tmpDir,
+      paste(
+        paste(
+          "tmp",
+          "calibration",
+          "scenario",
+          scenarioId,
+          sep = "_"
+        ),
+        "csv",
+        sep = "."
+      )
+    )
+  )
+  
+  readr::write_csv(
+    evaluation$calibration,
+    file.path(
+      tmpDir,
+      paste(
+        paste(
+          "tmp",
+          "concordance",
+          "scenario",
+          scenarioId,
+          sep = "_"
+        ),
+        "csv",
+        sep = "."
+      )
+    )
+  )
+  
+  readr::write_csv(
+    evaluation$adaptiveModel,
+    file.path(
+      tmpDir,
+      paste(
+        paste(
+          "tmp",
+          "adaptiveModel",
+          "scenario",
+          scenarioId,
+          sep = "_"
+        ),
+        "csv",
+        sep = "."
+      )
+    )
+  )
+  
+  return("done")
+  
+}
+
+
+mergeTmpFiles <- function(metric, tmpDir, saveDir) {
+  files <- list.files(
+    path=tmpDir,
+    full.names = TRUE, 
+    pattern = paste(
+      "tmp",
+      metric,
[…]

### 2022-04-12 extras/shiny/global.R

@@ -1,0 +1,67 @@
+library(tidyverse)
+rmse <- readr::read_csv(
+  "data/rmse.csv",
+  col_types = readr::cols(.default = "d")
+)
+discrimination <- readr::read_csv(
+  "data/discrimination.csv",
+  col_types = readr::cols(.default = "d")
+)
+calibration <- readr::read_csv(
+  "data/calibration.csv",
+  col_types = readr::cols(.default = "d")
+)
+
+analysisIds1 <- readr::read_csv(
+  "data/analysisIds.csv"
+) %>%
+  select(scenario, base, type, harm, sampleSize, auc)
+
+analysisIds2 <- readr::read_csv("data/analysisIdsInteractions.csv") %>%
+  select(scenario, base, type, harm, sampleSize, auc)
+
+analysisIds <- bind_rows(analysisIds1, analysisIds2)
+
+
+
+
+createPlot2 <- function(data) {
+  tmp <- data %>%
+    reshape2::melt()
+  levels(tmp$variable) <- case_when(
+    levels(tmp$variable) == "linear_predictor"          ~ "linear predictor",
+    levels(tmp$variable) == "constant_treatment_effect" ~ "constant treatment\neffect",
+    TRUE                                                ~ levels(tmp$variable)
+  )
+  plotly::plot_ly(
+    x = ~ variable,
+    y = ~ value,
+    color = ~ variable,
+    colors = c(
+      "#1b9e77",
+      "#d95f02",
+      "#7570b3",
+      "#e7298a",
+      "#66a61e",
+      "#e6ab02",
+      "#a6761d"
+    ),
+    data = tmp
+  ) %>%
+    plotly::add_trace(
+      type = "box",
+      hoverinfo = "y+q"
+    ) %>%
+    plotly::layout(
+      yaxis = list(
+        hoverformat = ".3f",
+        title = "",
+        gridcolor = "ffff"
+      ),
+      xaxis = list(
+          title = ""
+      ),
+      plot_bgcolor = '#e1e9f5'
+    )
+}
+

### 2022-04-12 extras/shiny/server.R

@@ -1,0 +1,203 @@
+shiny::shinyServer(
+  function(
+    input,
+    output,
+    session
+  ) {
+    
+    selectedVals                       <- shiny::reactiveValues()
+    selectedVals$base                  <- "absent"
+    selectedVals$type                  <- "constant"
+    selectedVals$nPatients             <- 4250
+    selectedVals$predictionPerformance <- .75
+    selectedVals$harm                  <- "absent"
+    
+    observeEvent(
+      input$base, 
+      {
+        req(
+          input$type, 
+          input$nPatients, 
+          input$predictionPerformance,
+          input$harm
+        )
+        
+        selectedVals$type                  <- input$type
+        selectedVals$nPatients             <- input$nPatients
+        selectedVals$predictionPerformance <- input$predictionPerformance
+        selectedVals$harm                  <- input$harm
+      }
+    )
+    
+    
+    
+    output$typeInput <- shiny::renderUI(
+      {
+        if (input$base != "interaction") {
+          shiny::selectInput(
+            inputId  = "type",
+            label    = "Type",
+            choices  = c(
+              "constant",
+              "linear-moderate",
+              "linear-high",
+              "quadratic-moderate",
+              "quadratic-high",
+              "non-monotonic"
+            ),
+            selected = selectedVals$effect
+          )
+        } else {
+          shiny::selectInput(
+            inputId  = "effect",
+            label    = "Effect",
+            choices  = c(
+              "weak",
+              "strong",
+              "mixed"
+            ),
+            selected = "weak"
+          )
+        }
+      }
+    )
+    
+    output$toggleNPatients <- shiny::renderUI(
+      {
+        if (input$base != "interaction") {
+          shiny::selectInput(
+            inputId  = "nPatients",
+            label    = "Number of patients",
+            choices  = c(
+              1063,
+              4250,
+              17000
+            ),
+            selected = selectedVals$nPatients
+          )
+        }
+      }
+    )
+    
+    output$togglePredictionPerformance <- shiny::renderUI(
+      {
+        if (input$base != "interaction") {
+          shiny::selectInput(
+            inputId  = "predictionPerformance",
+            label    = "Prediction model AUC",
+            choices  = c(
+              .65,
+              .75,
+              .85
+            ),
+            selected = selectedVals$predictionPerformance
+          )
+        }
+      }
+    )
+    
[…]

### 2022-04-12 extras/shiny/ui.R

@@ -1,0 +1,98 @@
+shiny::shinyUI(
+  shinydashboardPlus::dashboardPage(
+    skin                 = "black",
+    title                = "Simulations",
+    shinydashboard::dashboardHeader(
+      title = "Simulations"
+    ),
+    shinydashboard::dashboardSidebar(
+      shinydashboard::sidebarMenu(
+        id = "menu1",
+        shinydashboard::menuItem(
+          tabName = "home",
+          text    = "Home",
+          icon    = icon("home")
+        ),
+        shinydashboard::menuItem(
+          tabName = "settings",
+          text    = "Settings",
+          icon    = icon("cogs")
+        )
+      ),
+      shinydashboard::sidebarMenu(
+        id = "menu2",
+        shiny::conditionalPanel(
+          condition = "input.menu1 == 'settings'",
+          shiny::selectInput(
+            inputId  = "base",
+            label    = "Base",
+            choices  = c(
+              "absent",
+              "moderate",
+              "high",
+              "interaction"
+            ),
+            selected = "absent"
+          ),
+          shiny::uiOutput("typeInput"),
+          shiny::uiOutput("toggleNPatients"),
+          shiny::uiOutput("togglePredictionPerformance"),
+          shiny::selectInput(
+            inputId  = "harm",
+            label    = "Constant harm",
+            choices  = c(
+              "absent",
+              "moderate-positive",
+              "strong-positive",
+              "negative"
+            ),
+            selected = "absent"
+          )
+        )
+      )
+    ),
+    shinydashboard::dashboardBody(
+      shinydashboard::tabItems(
+        shinydashboard::tabItem(
+          tabName = "home",
+          shiny::includeHTML(
+            "html/abstract.html"
+          )
+        ),
+        shinydashboard::tabItem(
+          tabName = "settings",
+          shiny::tabsetPanel(
+            id = "results",
+            shiny::tabPanel(
+              title = "Description"
+            ),
+            shiny::tabPanel(
+              title = "RMSE",
+              plotly::plotlyOutput(
+                "rmsePlot",
+                height = "600px",
+                width = "1000px"
+              )
+            ),
+            shiny::tabPanel(
+              title = "Discrimination",
+              plotly::plotlyOutput(
+                "discriminationPlot",
+                height = "600px",
+                width = "1000px"
+              )
+            ),
+            shiny::tabPanel(
+              title = "Calibration",
+              plotly::plotlyOutput(
+                "calibrationPlot",
[…]

### 2022-04-13 submission/supplement.rmd
fix y-axis in shiny plots + add directory descriptions to readme + update formulas in supplement

@@
 \begin{equation}
-P(Y=1\given X=x,Z=z) = g(x^t\beta + \gamma z)
+P(Y=1\given X=x,Z=z) = g(x^t\beta + \delta_0 z)
 \end{equation}
@@
 
-where $f(\hat{lp}, z) = \hat{lp}+\hat{\gamma}z$, with $\hat{\gamma}$ the
+where $f(\hat{lp}, z) = \hat{lp}+\hat{\delta}_0z$, with $\hat{\delta}_0$ the
 estimated relative treatment effect (log odds ratio).
@@
 
-$$ f(\hat{lp}, z) = \gamma_0+\gamma_1z+\gamma_2\hat{lp}+\gamma_3z\hat{lp} $$
+$$ f(\hat{lp}, z) = \delta_0+\delta_1z+\delta_2\hat{lp}+\delta_3z\hat{lp} $$
 
@@
 
-$$ f(\hat{lp}, z) = \gamma_0 + \gamma_1z+zs(\hat{lp}) $$
+$$ f(\hat{lp}, z) = \delta_0 + \delta_1z+zs(\hat{lp}) $$
 where

### 2022-04-13 extras/shiny/global.R

@@ -1,23 +1,23 @@
 library(tidyverse)
 rmse <- readr::read_csv(
-  "data/rmse.csv",
+  "https://raw.githubusercontent.com/mi-erasmusmc/HteSimulationRCT/main/data/processed/rmse.csv",
   col_types = readr::cols(.default = "d")
 )
 discrimination <- readr::read_csv(
-  "data/discrimination.csv",
+  "https://raw.githubusercontent.com/mi-erasmusmc/HteSimulationRCT/main/data/processed/discrimination.csv",
   col_types = readr::cols(.default = "d")
 )
 calibration <- readr::read_csv(
-  "data/calibration.csv",
+  "https://raw.githubusercontent.com/mi-erasmusmc/HteSimulationRCT/main/data/processed/calibration.csv",
   col_types = readr::cols(.default = "d")
 )
 
 analysisIds1 <- readr::read_csv(
-  "data/analysisIds.csv"
+  "https://raw.githubusercontent.com/mi-erasmusmc/HteSimulationRCT/main/data/processed/analysisIds.csv"
 ) %>%
   select(scenario, base, type, harm, sampleSize, auc)
 
-analysisIds2 <- readr::read_csv("data/analysisIdsInteractions.csv") %>%
+analysisIds2 <- readr::read_csv("https://raw.githubusercontent.com/mi-erasmusmc/HteSimulationRCT/main/data/processed/analysisIdsInteractions.csv") %>%
   select(scenario, base, type, harm, sampleSize, auc)
 
 analysisIds <- bind_rows(analysisIds1, analysisIds2)
@@ -61,7 +61,7 @@ createPlot2 <- function(data) {
       xaxis = list(
           title = ""
       ),
-      plot_bgcolor = '#e1e9f5'
+      plot_bgcolor = '#edf0f5'
     )
 }
 

### 2022-04-13 extras/shiny/server.R

@@ -164,12 +164,12 @@ shiny::shinyServer(
         tmp <- rmseSubset()
         tmp %>%
           select(-scenarioId) %>%
-          createPlot2()
-          # plotly::layout(
-          #   yaxis = list(
-          #     range = c(0, .15)
-          #   )
-          # )
+          createPlot2() %>%
+          plotly::layout(
+            yaxis = list(
+              range = c(0, .15)
+            )
+          )
       }
     )
     

### 2022-04-15 submission/manuscript.rmd
update manuscript format for submission

@@
 abstract: |
-  **Objective**: To compare different risk-based methods for optimal prediction
-  of treatment effects. **Study Design and Setting**: We simulated RCT data
-  using diverse assumptions for the average treatment effect, a baseline
-  prognostic index of risk (PI), the shape of its interaction with treatment
-  (none, linear, quadratic or non-monotonic), and the magnitude of
-  treatment-related harms (none or constant independent of the PI). We predicted
-  absolute benefit using: models with a constant relative treatment effect;
-  stratification in quarters of the PI; models including a linear interaction of
-  treatment with the PI; models including an interaction of treatment with a
-  restricted cubic spline (RCS) transformation of the PI; an adaptive approach
-  using Akaike’s Information Criterion. We evaluated predictive performance
-  using root mean squared error and measures of discrimination and calibration
-  for benefit. **Results**: The linear-interaction model displayed optimal or
-  close-to-optimal performance across many simulation scenarios with moderate
-  sample size (N=4,250; ~800 events). The RCS-model was optimal for strong
-  non-linear deviations from a constant treatment effect, particularly when
-  sample size was larger (N=17,000). The adaptive approach also required larger
-  sample sizes. These findings were illustrated in the GUSTO-I
-  trial. **Conclusion**: An interaction between baseline risk and treatment
-  assignment should be considered to improve treatment effect predictions.
 author:
-  - name: Alexandros Rekkas
-    affiliation: 1
-    correspondingauthor: true
-  - name: Peter R. Rijnbeek
-    affiliation: 1
-  - name: David M. Kent
-    affiliation: 2
-  - name: Ewout W. Steyerberg
-    affiliation: 3
-  - name: David van Klaveren
-    affiliation: 4
-address:
-  - code: 1
-    address: Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
-  - code: 2
-    address: Predictive Analytics and Comparative Effectiveness Center, Institute for Clinical Research and Health Policy Studies, Tufts Medical Center, Boston, Massachusetts, USA
-  - code: 3
-    address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
-  - code: 4
-    address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
+  - Alexandros Rekkas$^1$,
+  - Peter R. Rijnbeek$^1$,
+  - David M. Kent$^2$,
+  - Ewout W. Steyerberg$^3$,
+  - David van Klaveren$^4$
 numbersections: true
 linenumbers: false
-journal: arXiv.org
-keywords:
-  - treatment effect heterogeneity
-  - absolute benefit
-  - prediction models
 output:
-  rticles::elsevier_article:
-    keep_tex: true
+  bookdown::pdf_document2: default
   bookdown::word_document2:
@@
   - \usepackage{setspace}
+  - \usepackage[left, pagewise]{lineno}
+  - \usepackage{caption}
   - \usepackage{amsmath}
@@
   - \usepackage{bm}
-  - \usepackage{caption}
   - \usepackage{booktabs}
@@
 
+\thispagestyle{empty}
+\vspace{8mm}
+
+$^1$ Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
+
+$^2$ Predictive Analytics and Comparative Effectiveness Center, Institute for Clinical Research and Health Policy Studies, Tufts Medical Center, Boston, Massachusetts, USA
+
+$^3$ Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
+
+$^4$ Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
+
+\vspace{10mm}
+**Corresponding author**
+\singlespacing 
+Alexandros Rekkas, MSc
+
+Department of Medical Informatics
+
+Erasmus University Medical Center
+
+3000 CA Rotterdam, P.O. Box 2040
+
+Email: a.rekkas@erasmusmc.nl
+\onehalfspacing
+
+\vspace{10mm}
+**Funding**
+
+This work has been performed in the European Health Data and
+Evidence Network (EHDEN) project. This project has received funding from the
+Innovative Medicines Initiative 2 Joint Undertaking (JU) under grant agreement
+No 806968. The JU receives support from the European Union’s Horizon 2020
+research and innovation programme and EFPIA.
+
+\newpage
+\newpage
+
+# Abstract {-}
+**Objective**: To compare different risk-based methods for optimal prediction
+of treatment effects. **Study Design and Setting**: We simulated RCT data
+using diverse assumptions for the average treatment effect, a baseline
+prognostic index of risk (PI), the shape of its interaction with treatment
+(none, linear, quadratic or non-monotonic), and the magnitude of
+treatment-related harms (none or constant independent of the PI). We predicted
+absolute benefit using: models with a constant relative treatment effect;
+stratification in quarters of the PI; models including a linear interaction of
+treatment with the PI; models including an interaction of treatment with a
+restricted cubic spline (RCS) transformation of the PI; an adaptive approach
+using Akaike’s Information Criterion. We evaluated predictive performance
+using root mean squared error and measures of discrimination and calibration
+for benefit. **Results**: The linear-interaction model displayed optimal or
+close-to-optimal performance across many simulation scenarios with moderate
+sample size (N=4,250; ~785 events). The RCS-model was optimal for strong
+non-linear deviations from a constant treatment effect, particularly when
+sample size was larger (N=17,000). The adaptive approach also required larger
+sample sizes. These findings were illustrated in the GUSTO-I
+trial. **Conclusion**: An interaction between baseline risk and treatment
+assignment should be considered to improve treatment effect predictions.
+
+\vspace{10mm}
+
+**Keywords**: treatment effect heterogeneity, absolute benefit, prediction models
+
+\newpage
+
 \doublespacing
-\linenumbers
 
@@
 Risk-modeling approaches to predictive HTE analyses provide a viable option in
-the absence of well-established treatment effect modifier [@Kent2019;
+the absence of well-established treatment effect modifiers [@Kent2019;
 @PathEnE]. In simulations, modeling treatment-covariate interactions, often led
@@
 $$\tau(x) = E\{Y(0) - Y(1)|X=x\}$$ 
-Assuming that $(Z, X, Y)$ is a random sample from the target population and that
-$\big(Y(0), Y(1)\big)\indep Z|X$, as we are in the RCT setting, we can predict
-CATE from
+Assuming that $\big(Y(0), Y(1)\big)\indep Z|X$, as we are in the RCT setting, we
+can predict CATE from
 \begin{align*}
@@
 These settings resulted in a simulation study of 648 scenarios covering the
-observed HTE in 32 large trials as well as many other potential variations of
+HTE observed in 32 large trials as well as many other potential variations of
 risk-based treatment effect (Supplement, Sections 2 and 3) [@Kent2016].
@@
 slightly outperformed the linear interaction model
-(Median RMSE: linear interaction `r median(rmseNonMonotonic$linear_predictor)`;
+(median RMSE: linear interaction `r median(rmseNonMonotonic$linear_predictor)`;
 RCS-3 `r median(rmseNonMonotonic$rcs_3_knots)`; Figure \ref{fig:rmsebase} panel D).
@@
 the optimal performance of RCS-3 was more pronounced
-(Median RMSE: linear interaction `r median(rmseNonMonotonicHarm$linear_predictor)`;
+(median RMSE: linear interaction `r median(rmseNonMonotonicHarm$linear_predictor)`;
 RCS-3 `r median(rmseNonMonotonicHarm$rcs_3_knots)`; Figure \ref{fig:rmsebase} panel D).
@@
 
-Increasing the sample size to `r 17000` favored RCS-3 the most. The difference in
-performance with the linear interaction approach was more limited in settings
-with a constant treatment effect 
-(Median RMSE: linear interaction `r median(rmseBaseCaseSampleSize$linear_predictor)`;
+Increasing the sample size to `r 17000` favored RCS-3 the most (Figure
+\ref{fig:rmsesamplesize}). The difference in performance with the linear
+interaction approach was more limited in settings with a constant treatment
+effect
+(median RMSE: linear interaction `r median(rmseBaseCaseSampleSize$linear_predictor)`;
 RCS-3 `r median(rmseBaseCaseSampleSize$rcs_3_knots)`) and with a true linear interaction
-(Median RMSE: linear interaction `r median(rmseLinearInteractionSampleSize$linear_predictor)`;
-RCS-3 `r median(rmseLinearInteractionSampleSize$rcs_3_knots)`).
+(median RMSE: linear interaction `r median(rmseLinearInteractionSampleSize$linear_predictor)`;
+RCS-3 `r median(rmseLinearInteractionSampleSize$rcs_3_knots)`)
 and more emphasized in settings with strong quadratic deviations 
-(Median RMSE: linear interaction `r median(rmseQuadraticSampleSize$linear_predictor)`;
+(median RMSE: linear interaction `r median(rmseQuadraticSampleSize$linear_predictor)`;
 RCS-3 `r median(rmseQuadraticSampleSize$rcs_3_knots)`) and non-monotonic deviations
-(Median RMSE: linear interaction `r median(rmseNonMonotonicSampleSize$linear_predictor)`;
+(median RMSE: linear interaction `r median(rmseNonMonotonicSampleSize$linear_predictor)`;
 RCS-3 `r median(rmseNonMonotonicSampleSize$rcs_3_knots)`).
@@
 `r median(rmseLinearInteractionAuc$linear_predictor)` for the linear interaction
-model). Similar to the base case scenario the adaptive approach wrongly selected
-the constant treatment effect model
+model; Figure \ref{fig:rmseauc}). Similar to the base case scenario the adaptive
+approach wrongly selected the constant treatment effect model
 (`r adaptiveSelections %>% filter(scenarioId == 297) %>% pull(treatment) * 100`\%
@@
 The results from all individual scenarios can be explored online at
-[https://arekkas.shinyapps.io/simulation_viewer/](https://arekkas.shinyapps.io/simulation_viewer/). Additionally,
+[https://mi-erasmusmc.shinyapps.io/HteSimulationRCT/](https://mi-erasmusmc.shinyapps.io/HteSimulationRCT/). Additionally,
 all the code for the simulations can be found at
-[https://github.com/rekkasa/arekkas_HteSimulation_XXXX_2021](https://github.com/rekkasa/arekkas_HteSimulation_XXXX_2021)
+[https://github.com/mi-erasmusmc/HteSimulationRCT](https://github.com/mi-erasmusmc/HteSimulationRCT)
 
@@
 risk in the majority of the simulation scenarios. We attempted to expand our
-scenarios by considering constant moderate and strong treatment-related harms,
+scenarios by considering moderate and strong constant treatment-related harms,
 applied on the absolute scale, in line with previous work [@Glasziou1995]. In a

### 2022-08-31 code/CreateBootstrapData.R
add bootstrap confidence intervals and rebuild

@@ -1,0 +1,152 @@
+#!/usr/bin/env Rscript
+
+# ===================================================
+# Description:
+#   Generates a dataframe with bootstrap confidence
+#   intervals.
+# Input:
+#   - number of threads for parallelization
+# Output:
+#   - data/processed/bootstrapData.csv
+# Depends:
+#   - data/raw/gusto.rda
+# ===================================================
+
+library(tidyverse)
+library(rms)
+library(SmoothHte)
+library(ParallelLogger)
+
+set.seed(19910930)
+
+
+load("data/raw/gusto.rda")
+gusto <- gusto %>%
+  tibble() %>%
+  filter(tx != "SK+tPA") %>%
+  rename(
+    "outcome" = "day30",
+    "treatment" = "tpa"
+  )
+
+prediction <- lrm(
+  outcome ~ treatment + age + Killip + pmin(sysbp, 120) + lsp(pulse, 50) + pmi + miloc,
+  data = gusto,
+  maxit = 99
+)
+
+riskLinearPredictor <- predict(
+  prediction,
+  newdata = gusto %>%
+    mutate(treatment = 0) %>%
+    data.frame()
+)
+
+gusto <- gusto %>%
+  mutate(
+    riskLinearPredictor = riskLinearPredictor
+  )
+
+bootstrapConfidenceInterval <- function(x, data) {
+  set.seed(19900930 + x)
+  originalSampleSize <- nrow(data)
+  bootRows <- sample(
+    x = 1:originalSampleSize,
+    size = originalSampleSize,
+    replace = TRUE
+  )
+  bootData <- data[bootRows, ]
+  prediction <- lrm(
+    outcome ~ treatment + age + Killip + pmin(sysbp, 120) + lsp(pulse, 50) + pmi + miloc,
+    data = bootData,
+    maxit = 99
+  )
+
+  riskLinearPredictor <- predict(
+    prediction,
+    newdata = bootData %>% mutate(treatment = 0) %>% data.frame()
+  )
+
+  bootData <- bootData %>%
+    mutate(
+      riskLinearPredictor = riskLinearPredictor
+    )
+  constantModel <- fitModelBasedHte(
+    data     = bootData,
+    settings = createModelBasedSettings(type = "treatment")
+  )
+  linearModel <- fitModelBasedHte(
+    data     = bootData,
+    settings = createModelBasedSettings()
+  )
+
[…]

### 2022-08-31 code/GustoPlot.R

@@ -1,12 +1,23 @@
 #!/usr/bin/env Rscript
 
+# ===================================================
+# Description:
+#   Generates the plot for gusto trial.
+# Input:
+# Output:
+#   - figures/gusto.tiff
+# Depends:
+#   - data/raw/gusto.rda
+#   - data/processed/bootstrapData.csv
+# ===================================================
+
 library(tidyverse)
 library(rms)
 library(SmoothHte)
 
-load("data/raw/gusto.rda")
-maxRisk <- .3      # the maximum risk to be plotted
+maxRisk <- .25
 
+load("data/raw/gusto.rda")
 gusto <- gusto %>%
   tibble() %>%
   filter(tx != "SK+tPA") %>%
@@ -23,13 +34,16 @@ prediction <- lrm(
 
 riskLinearPredictor <- predict(
   prediction,
-  newdata = gusto %>% mutate(treatment = 0) %>% data.frame()
+  newdata = gusto %>%
+    mutate(treatment = 0) %>%
+    data.frame()
 )
 
 gusto <- gusto %>%
   mutate(
     riskLinearPredictor = riskLinearPredictor
   )
+bootstrapData <- readr::read_csv("data/processed/bootstrapData.csv")
 
 constantModel <- fitModelBasedHte(
   data     = gusto, 
@@ -51,73 +65,13 @@ rcs3Model <- fitRcsHte(
   settings = createRcsSettings()
 )
 
-plot <- ggplot() +
-  geom_pointrange(
-    data = stratified$data,
-    aes(
-      x     = meanRisk, 
-      y     = estimate, 
-      ymin  = lower, 
-      ymax  = upper,
-      color = "Stratified"
-    ),
-    key_glyph = "rect"
-  ) +
-  stat_function(
-    data = data.frame(
-      x     = c(0, maxRisk), 
-      label = "Constant treatment effect"
-    ),
-    aes(
-      x     = x, 
-      color = label,
-      linetype = label
-    ),
-    size = .8,
-    alpha = .8,
-    fun  = predictBenefitModelBasedHte,
-    args = list(
-      modelBasedFit = constantModel
-    )
-  ) +
-  stat_function(
-    data = data.frame(
-      x = c(0, maxRisk), 
-      label = "Linear interaction"
-    ),
-    aes(
-      x     = x,
-      linetype = label,
-      color = label
-    ),
-    size = .65,
-    fun  = predictBenefitModelBasedHte,
-    args = list(
[…]

### 2022-09-04 submission/reviewer_response.rmd
drafting changes based on reviewer comments

---
title: "Response to reviewers"
output:
  bookdown::pdf_document2: default
geometry: margin=1.0in
toc: true
font-size: 12pt
editor_options: 
  chunk_output_type: console
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage[left, pagewise]{lineno}
  - \usepackage{caption}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \newcommand{\indep}{\perp \!\!\! \perp}
bibliography: references.bib
csl: jce.csl
---

# Reviewer 1 {-}
"Individualized treatment effect was predicted best by modeling baseline risk in
interaction with treatment assignment"
Comments to the Authors:

I read the manuscript with interest, however, the authors first need to address
the following concerns.

- Page 2.  Line 6. Need to clarify what it means by baseline risk, as well as
what it means by the “reference.”  Line 10. Related to the comment above, can
you provide a sentence or two briefly describing what risk-modeling approaches
are?  Line 12. Need to clarify what it means by “absolute benefit.”
Line 13. Not clear, from the written sentence, which methods (whether they are
risk-based methods or the method based on modeling treatment effect-modifiers)
lead to the weaker discrimination of benefit in the presence of true effect
modifiers.  Line 18. The estimates refer to estimates of which parameters? Need
to clarify. Also, please indicate why these estimates may not be applied to
individual patients. \textcolor{green}{[DONE]}

- Page 4.  Line 1. Not clear what it means by the "AUC value of 0.75." Need to
explain AUC. And after all, what is the "true" value of $\beta$ that was used in the
simulations?  Line 6. What is c? And which term in the equation is the term that
corresponds to the constant treatment effect? What are the values of vs used in
the simulations?  Line 11. What do you mean by “standard” alpha? what is
"standard"? What is "effect"? Does it refer to the effect on the approaches'
performance? What do you mean by the effect of risk model discriminative
ability? How can risk model discriminative ability have any effects? Please
clarify. \textcolor{green}{[DONE]}

- Page 5.  Line 14. Please clarify how $\delta_0$, $\delta_1$ and $\delta_3$ are
  estimated. \textcolor{green}{[DONE]}

- Page 6.  Line 8. Please describe in a sentence what "ICI for benefit" is.
Line17. Please comment on how those 6 predictors were selected (is it also
selected based on AIC?). \textcolor{green}{[DONE]}

-  Page 7.  Line 14. Please describe how the "absolute benefit predictions" were
evaluated (please present the evaluation metric). \textcolor{green}{[DONE]}

- Page 8. In the figures, I think for each panel you should label the X-axis
(i.e., the methods under consideration). At first, it was difficult to
understand what was displayed. \textcolor{green}{[DONE]}

- Page 13. In Figure 4, again, what is the C-statistic for benefit? Need an
explicit expression or a clear description (both in figure caption and in the
text). \textcolor{green}{[DONE]}

- Page 16. In Figure 6, please also display the model-based confidence bands for
all the methods. Also please clarify how the predicted benefit was
computed. Does it just mean the treatment effect? If so, how could the "constant
treatment effect" method have a non-constant treatment effect as a function of
the baseline risk? Please clarify. \textcolor{green}{[DONE]}

- Page 18.  Line 21. Generally, in the estimation, how do you distinguish the
treatment-related harm from the actual treatment effect? \textcolor{red}{[UNRESOLVED]}

- Page 19.  Line 1. As discussed in the Discussion, the authors only considered
the risk-based methods. Along this line, I think the title of the manuscript is,
to some extent, misleading, because it gives an impression that the modeling
baseline risk-and-treatment interaction is a “globally best method” for the
individualized treatment effect estimation, which is not likely to be true.
Line 2. Why only tree-based approaches were cited? There is a large volume of
treatment effect modeling methods, which I think should be cited as well, if
tree-based approaches are to be cited.  Line 8. I don’t think what is stated in
line 8 is true when the "true" data model is defined based on separate
treatment-by-covariate interaction terms, in addition to the main effect
terms. Furthermore, there are many methods that avoid overfitting through
regularization and simultaneous treatment effect modifier selection. Authors
should discuss this. \textcolor{red}{[UNRESOLVED]}

# Reviewer 2 {-}

I reviewed the same manuscript last month. It seems that the manuscript is not
revised and is the same as the previous version and did not address any of the
comments. Please see my comments made in the previous round of review. Would you
please check if this attached manuscript is a revised version ? Thank you.

# Reviewer 3 {-}

The content of this paper is methodologically sound and the results
are consistent with what the literature has shown (namely that more complex
models are often outperformed by simpler ones in small sample and weaker signal
scenarios) and the writing is generally adequate. However, as written, the paper
does not focus on the significance of this methodology and the simulation
studies in terms of guiding clinical care (see Comment 1 for more details); this
was the main weakness of this paper.

1. The introduction will benefit from more emphasis on the practical
   significance of modeling baseline risk and treatment effects. In particular,
   in which real-life scenarios was this type of analysis useful and what
   happened when the baseline risk and its interaction with treatment was
   neglected in favor of a simpler approach? What type of analyses are typically
   done in such scenarios and what does the approach presented in this
   manuscript offer on top of the usual analysis methods?

2. During the application of restricted cubic splines, was there an assumption
   of interaction between risk and treatment effects as well? For example, were
   the restricted cubic splines fit separately to each treatment arm?

3. It would also be helpful to discuss the results (particularly the fact that
   the simpler model--linear interaction--outperforms RCS-3 in smaller sample
   size scenarios and weaker signals, even when there is nonlinearity in the
   true underlying relationship, as well as the fact that RCS-3 outperforms
   RCS-4 and RCS-5 in many scenarios) in terms of the bias-variance tradeoff. In
   particular, the restricted cubic splines-based methods introduce additional
   degrees of freedom and thus increase the variance, despite being expected to
   reduce the bias.

# Reviewer 4 {-}

This paper presents the results from a very large simulation study to
evaluate a number of fairly simple statistical approaches to estimate individual
treatment effects using data from a RCT. The paper is well written and the
results are presented objectively. The results show essentially what you might
expect, that if the data generating method is more complicated then flexible
methods perform best, while if the data generation method is more simple then
simpler methods do best. Furthermore, with larger sample size flexible methods
tend to do better. The authors end up essentially recommending an approach that
has a linear interaction term, or one with a cubic spline. A couple of the
comparators are even more simple than these (constant treatment effect and
stratification by quartiles), and it is no surprise that these naïve method are
not good in the simulations. So if the point of the paper is to discourage use
of these naïve methods, then that is a valuable contribution.  The big
limitation of the whole study is that the baseline risk (i.e. the linear
combination of the X variables from the control arm of the RCT) is assumed to be
the driver of the individual treatment effect. The data is generated with this
assumption, and all the estimation methods make this assumption. The authors
totally acknowledge this limitation. But I would have thought that in the
majority of realistic situations the linear combination of X’s for the main
effect would have been different from the linear combination of X’s in the
interaction with treatment. In this case the authors conclusion that linear
interaction and RCS3 are good, would likely not hold up. The authors do
reference a few articles that use tree-based methods (there are many many more
and lots of non tree-based approaches too), and correctly point out their
potential for overfitting. But regularization and cross-validation approaches
can do a pretty good job at limiting the overfitting. So I think I would favor
tree-based (or other more modern approaches), unless there is some clinical or
scientific rationale to believe that the control group risk is the driver of the
treatment effect heterogeneity.  A couple of others things that should be
mentioned as caveats. In the simulated data the X variables are made independent
of each other. If they were made to be correlated, that would probably be more
realistic. Do you have any information on how this might change the findings, in
general terms.  Larger sample sizes also opens up the possibility of
generalizing the linear additive assumption for the control group risk. For
example, this part of the model could now include non-linear terms in X and
interactions between X’s.

### 2022-09-04 submission/manuscript_tracked.rmd
drafting changes based on reviewer comments

--- 2022-04-15 submission/manuscript.rmd
@@
 title: |
- Individualized treatment effect was predicted best by modeling 
+ Individualized treatment effect was predicted best by modelin
  baseline risk in interaction with treatment assignment
@@
 
-```{r, echo=FALSE, message=FALSE}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 library(tidyverse)
@@
 development of models predicting either individualized effects or which of two
-(or more) treatments is better for an individual [@Varadhan2013]. In prior work,
-we divided such methods in three broader categories based on the reference class
-used for defining patient similarity when making individualized predictions or
-recommendations [@Rekkas2020]. Risk-modeling approaches use prediction of
-baseline risk as the reference; treatment effect modeling approaches also model
-treatment-covariate interactions, in addition to risk factors; optimal treatment
-regime approaches focus on developing treatment assignment rules and rely
-heavily on modeling treatment effect modifiers.
+(or more) treatments is better for an individual
+
+:::{custom-style="bluefont"}
+with regard to a specific outcome of interest
+:::
+
+[@Varadhan2013]. In prior work, we divided such
+methods in three broader categories based on the reference
+class
+
+:::{custom-style="bluefont"}
+, i.e. the set of covariates} used for defining patient
+:::
+
+similarity when making individualized predictions or recommendations
+[@Rekkas2020]. ~~Risk-modeling approaches use prediction of baseline risk as the
+reference;~~
+
+:::{custom-style="bluefont"}
+Risk modeling methods predict similar treatment benefits for patients with
+similar baseline outcome risk, i.e. the probability of a certain patient
+experiencing the outcome of interest under the control;
+:::
+
+treatment effect modeling approaches also model treatment-covariate
+interactions, in addition to risk factors; optimal treatment regime approaches
+focus on developing treatment assignment rules and rely heavily on modeling
+treatment effect modifiers.
 
@@
 the absence of well-established treatment effect modifiers [@Kent2019;
-@PathEnE]. In simulations, modeling treatment-covariate interactions, often led
-to miscalibrated predictions of absolute benefit, contrary to risk-based
-methods, despite their weaker discrimination of benefit in the presence of true
-effect modifiers [@vanKlaveren2019]. Most often, risk-modeling approaches are
-carried out in two steps: first a risk prediction model is developed externally
-or internally on the entire RCT population, “blinded” to treatment; then the RCT
-population is stratified using this prediction model to evaluate risk-based
-treatment effect variation [@Kent2010]. This approach identified substantial
-absolute treatment effect differences between low-risk and high-risk patients in
-a re-analysis of 32 large trials [@Kent2016]. However, even though estimates at
-the risk subgroup level may be accurate, these estimates may not apply to
-individual patients.
+@PathEnE].
+
+:::{custom-style="bluefont"}
+Risk modeling approaches are not new [@Rekkas2020] and are quite intuitive to
+practitioners. Often medical guidelines rely on a risk stratified approach to
+target treatments to different patients. In addition, re-analyses of studies
+that only looked at overall results using risk stratification often result to
+important insight on how treatment effects varied for diffrent patients. For
+example, a risk stratified analysis of patients with acute myocardial infarction
+(MI) based on the Thrombolysis in Myocardial Infarction (TIMI) risk score found
+no benefit for patients who underwent primary angioplasty compared to
+fibrynolysis. However, there was a significant benefit for patients with a high
+TIMI score [@Thune2005]. Infants at lower risk of bronchopulmonary dysplasia
+benefit relatively more from vitamin A therapy than infants at higher risk
+[@Rysavy2021]. Finally, higher risk prediabetic patients benefit relatively more
+from metformin than lower risk patients [@Sussman2015].
+:::
+
+:::{custom-style="bluefont"}
+They result in increased power for detecting HTE, as they
+evaluate interactions of treatment with a summary score, i.e. baseline outcome
+risk, and, thus, avoid modeling treatment-covariate interactions.
+:::
+
+In simulations, modeling treatment-covariate interactions, often led to
+miscalibrated predictions of ~~absolute benefit~~
+
+:::{custom-style="bluefont"}
+benefit on
+the absolute scale (risk difference), ~~contrary~~ compared
+:::
+
+to risk-modeling methods~~, despite their weaker discrimination of benefit in
+the presence of true effect modifiers~~
+[@vanKlaveren2019].
+
+:::{custom-style="bluefont"}
+However, in the presence of true
+treatment-covariate interactions, risk modeling methods performed worse in
+separating lower from higher benefit patients compared to treatment effect
+modeling methods.
+:::
+
+Most often, risk-modeling approaches are carried out in two
+steps: first a risk prediction model is developed externally or internally on
+the entire RCT population, “blinded” to treatment; then the RCT population is
+stratified using this prediction model to evaluate risk-based treatment effect
+variation [@Kent2010]. This approach identified substantial absolute treatment
+effect differences between low-risk and high-risk patients in a re-analysis of
+32 large trials [@Kent2016]. However, even though
+
+:::{custom-style="bluefont"}
+treatment effect estimates at the risk subgroup level may be accurate, these
+estimates may not apply to individual patients,{as homogeneity of treatment
+effects is assumed within risk strata. With stronger treatment effect
+heterogeneity (larger variability between estimated risk-subgroup treatment
+effects), predicted treatment effects for patients with similar predicted risk
+may be quite different, if they end up in different risk strata.
+:::
 
@@
 of predicting individualized effects of treatment for acute myocardial
-infarction (MI) in a large RCT.
+infarction ~~(MI)~~in a large RCT.
 
@@
 with true probabilities following a logistic regression model including all
-baseline covariates, i.e.
-$P(Y(0)=1\given X=x) = \text{expit}(lp_0) = e^{lp_0}/(1+e^{lp_0})$,
-with $lp_0=lp_0(x)=x^t\beta$.
-In the base scenarios coefficient values $\beta$ were
-such, that the AUC of the logistic regression model was 0.75 and the event rate
-in the control arm was $20\%$.
+baseline covariates, i.e.  $P(Y(0)=1\given X=x) = \text{expit}(lp_0) =
+e^{lp_0}/(1+e^{lp_0})$, with $lp_0=lp_0(x)=x^t\beta$.  In the base scenarios
+coefficient values $\beta$ were such, that the ~~AUC of the logistic regression
+model was 0.75 and the event rate in the control arm was
+$20\%$.~~
+
+:::{custom-style="bluefont"}
+the control event rate was 20\% and the discriminative
+ability of the true prediction model measured using Harrell's c-statistic was
+0.75. The c-statistic represents the probability that for a randomly selected
+discordant pair from the sample (patients with different outcomes) the
+prediction model assigns larger risk to the patient with the worse outcome. For
+the simulations this was achieved by selecting $\beta$ values such that the true
+prediction model would achieve a c-statistic of 0.75 in a simulated control arm
+with 1,000,000 patients. In the base case scenario we achieved that by setting
+$\beta=(-2.08, 0.49,\dots,0.49)^t$.
+:::
 
@@
 treatment effect. We then introduced linear, quadratic and non-monotonic
-deviations from constant treatment effects using:
-$$lp_1 = \gamma_2(lp_0-c)^2 + \gamma_1(lp_0-c) + \gamma_0, $$
-where $lp_1$ is the true linear predictor in the treatment arm, so that
-$P(Y(1)=1\given X=x) = \text{expit}(lp_1)$.
-Finally, we incorporated constant absolute harms for all treated patients, such
-that
-$P(Y(1)=1|X=x) = \text{expit}(lp_1) + \text{harm}$.
+deviations from constant treatment effects using: $$lp_1 = \gamma_0 +
+\gamma_1(lp_0-c) + \gamma_2(lp_0-c)^2, $$ where $lp_1$ is the true linear
+predictor in the treatment arm, so that $P(Y(1)=1\given X=x) =
+\text{expit}(lp_1)$,
+
+:::{custom-style="bluefont"}
+$\gamma=(\gamma_0, \gamma_1, \gamma_2)^t$
+controls the shape of evolution of treatment effect as a function of baseline
+risk (type and strength of deviations from the constant treatment effect
+setting), while $c$ allows us to shift the proposed shape function to achieve
+the desired overall event rates. For example, to simulate a constant treatment
+effect with $\text{OR}=0.8$ we would set $\gamma=(\log(0.8), 1, 0)^t$ and
+$c=0$.
+:::
+
+Finally, we incorporated constant absolute harms for all treated
+patients, such that $P(Y(1)=1|X=x) = \text{expit}(lp_1) + \text{harm}$.
 
 The sample size for the base scenarios was set to 4,250 (80\% power for the
-detection of a marginal OR of 0.8 with the standard alpha of 5\%). We evaluated
-the effect of smaller or larger sample sizes of 1,063 and 17,000,
-respectively. We also evaluated the effect of risk model discriminative ability,
-adjusting the baseline covariate coefficients, such that the AUC of the
-regression model in the control arm was 0.65 and 0.85, respectively.
+detection of a marginal OR of 0.8 ~~with the standard alpha of 5\%~~
+
+:::{custom-style="bluefont"}
+using a 5\% significance level for the Wald test for the treatment coefficient
+in a logistic regression model).
+:::
+
+We evaluated the
+~~effect~~
+
+:::{custom-style="bluefont"}
+impact
+:::
+
+of smaller or larger sample sizes of 1,063
+and 17,000, respectively. We also evaluated the ~~effect~~
+
+:::{custom-style="bluefont"}
+impact
+:::
+
+of risk model discriminative ability, adjusting the
+baseline covariate coefficients, such that the AUC of the regression model in
+the control arm was 0.65 and 0.85, respectively.
 
@@
 
-Second, we considered a model which assumes *constant relative treatment effect*
-(constant odds ratio). Hence, absolute benefit is predicted from
-$\tau(x;\hat{\beta}) = \text{expit}(\hat{lp}_0) - \text{expit}(\hat{lp}_0+\delta_0)$, where
-$\delta_0$ is the log of the assumed constant odds ratio and
-$\hat{lp}_0 = \hat{lp}_0(x;\hat{\beta}) = x^t\hat{\beta}$
-the linear predictor of the estimated baseline risk model.
+Second, we ~~considered~~
 
-Third, we considered a logistic regression model including treatment, the
-prognostic index, and their linear interaction. Absolute benefit is then
-estimated from
-$\tau(x;\hat{\beta}) = \text{expit}(\delta_0+\delta_1\hat{lp}_0) - \text{expit}(\delta_0+\delta_2+(\delta_1+\delta_3)\hat{lp}_0)$
-We will refer to this method as the *linear interaction* approach.
+:::{custom-style="bluefont"}
+fitted
+:::
+
+a
+
+:::{custom-style="bluefont"}
+logistic regression 
+:::
+
+model which assumes *constant relative treatment effect* (constant
+odds ratio)
+
+:::{custom-style="bluefont"}
+, that is, $P(Y=1\given X=x, Z=z;\hat{\beta}) =
+\text{expit}(\hat{lp}_0 + \delta_0z)$.
+:::
+
+Hence, absolute benefit is predicted
+from $\tau(x;\hat{\beta}) = \text{expit}(\hat{lp}_0) -
+\text{expit}(\hat{lp}_0+\delta_0)$, where $\delta_0$ is the log of the assumed
+constant odds ratio and $\hat{lp}_0 = \hat{lp}_0(x;\hat{\beta}) =
+x^t\hat{\beta}$ the linear predictor of the estimated baseline risk model.
+
+Third, we ~~considered~~
+
+:::{custom-style="bluefont"}
+fitted
+:::
+
+a logistic regression model
+including treatment, the prognostic index, and their linear interaction
+
+:::{custom-style="bluefont"}
+, that is, $P(Y=1\given X=x, Z=z;\hat{\beta}) =
+\text{expit}(\delta_0+\delta_1z+\delta_2
+\hat{lp}_0+\delta_3z\hat{lp}_0)$
+:::
+
+Absolute benefit is then estimated from
+$\tau(x;\hat{\beta}) = \text{expit}(\delta_0+\delta_1\hat{lp}_0) -
+\text{expit}(\delta_0 + \delta_1\hat{lp}_0 + \delta_2z+\delta_3z\hat{lp}_0)$. We
+will refer to this method as the *linear interaction* approach.
 
@@
 c-for-benefit and the integrated calibration index (ICI) for
-benefit (Supplement, Section 6) [@vanKlaveren2018].
+benefit ~~(Supplement, section 6)~~.
+
+:::{custom-style="bluefont"}
+The c-for-benefit represents the probability that from two randomly chosen
+matched patient pairs with unequal observed benefit, the pair with greater
+observed benefit also has a higher predicted benefit. To be able to calculate
+observed benefit, patients in each treatment arm are ranked based on their
+predicted benefit and then matched 1:1 across treatment arms. Observed treatment
+benefit is defined as the difference of observed outcomes between the untreated
+and the treated patient of each matched patient pair. Predicted benefit is
+defined as the average of predicted benefit within each matched patient pair
+[@vanKlaveren2018].
+:::
+
+:::{custom-style="bluefont"}
+We evaluated calibration in a similar manner, using the integrated calibration
+index (ICI) for benefit [@Austin2019]. The observed benefits are regressed on
+the predicted benefits using a locally weighted scatterplot smoother
+(loess). The ICI-for-benefit is the average absolute difference between
+predicted and smooth observed benefit. Values closer to represent better
+calibration.
+:::
 
@@
 plasminogen activator (tPA) treatment and `r treatmentArms %>% filter(treatment
-== 0) %>% pull(n)` were randomized to streptokinase. The outcome of interest
-was 30-day mortality (total of `r sum(gusto$outcome)` events), recorded for all
-patients. In line with previous analyses [@Califf1997; @Steyerberg2000], we
-fitted a logistic regression model with 6 baseline covariates, i.e. age, Killip
-class, systolic blood pressure, heart rate, an indicator of previous MI, and the
-location of MI, to predict 30-day mortality risk (Supplement, Section 8).
+== 0) %>% pull(n)` were randomized to streptokinase. The outcome of interest was
+30-day mortality (total of `r sum(gusto$outcome)` events), recorded for all
+patients.
+
+:::{custom-style="bluefont"}
+This dataset has been used extensively in prior studies
+:::
+
+~~In line with previous analyses~~ [@Califf1997; @Steyerberg2000]
+
+:::{custom-style="bluefont"}
+Therefore, we used the same set of 7 covariates that was previously used to,
+:::
+
+~~we fitted~~
+
+:::{custom-style="bluefont"}
+fit
+:::
+
+a logistic regression
+model with ~~6~~
+
+:::{custom-style="bluefont"}
+the following
+:::
+
+baseline covariates: age, Killip class, systolic blood pressure, heart rate, an
+indicator of previous MI, and the location of MI, to predict 30-day mortality
+risk (Supplement, Section 8).
 
@@
 RCS-3 `r median(rmseNonMonotonicHarm$rcs_3_knots)`; Figure \ref{fig:rmsebase} panel D).
-A stronger average treatment effect (OR=0.5) led to larger absolute benefit
-predictions and consequently to larger RMSE for all approaches, but the relative
-differences between different approaches were similar to the base case scenario
+A stronger average treatment effect (OR=0.5)
+
+:::{custom-style="bluefont"}
+resulted in higher variability of the true treatment effects on the absolute
+scale (difference in true outcome probabilities between treatment arms)
+:::
+
+~~led to larger absolute benefit
+predictions~~ and consequently to larger RMSE for all approaches. ~~but~~
+
+:::{custom-style="bluefont"}
+However,
+:::
+
+the relative differences between different approaches were similar to the base case scenario
 (Supplement, Figure S10). 
-
 ```{r rmsebase, cache=TRUE, echo=FALSE, fig.cap="RMSE of the considered methods across 500 replications calculated from a simulated super-population of size 500,000.
@@
 
-```{r discrimination, cache=TRUE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated samples of size 4,250.
+```{r discrimination, cache=TRUE, echo=FALSE, fig.cap="Discrimination for benefit of the considered methods across 500 replications calculated in a simulated samples of size 4,250 using the c-statistic for benefit.
+The c-statistic for benefit represents the probability that from two randomly chosen matched patient pairs with unequal observed benefit, the pair with greater observed benefit also has a higher predicted benefit.
 True prediction AUC of 0.75.", fig.show="hold", out.width = '100%'}
@@
 predictions. Following our simulation results, RCS-4 and RCS-5 models were
-excluded. Finally, an adaptive approach with the 3 candidate models was applied.
+excluded. Finally, an adaptive approach with the 3 candidate models was
+applied.
 
-All considered methods provided similar fits, predicting increasing benefits for
-patients with higher baseline risk predictions, and followed the evolution of
-the stratified estimates closely (Figure \ref{fig:gusto}). The constant
-treatment effect model had somewhat lower AIC compared to the linear interaction
-model (AIC: 
+:::{custom-style="bluefont"}
+Predicted absolute benefit was derived as the difference of predicted acute MI
+risk between treatment arms, if all other predictors remained unchanged.
+:::
+
+All considered methods provided similar fits, predicting increasing
+
+:::{custom-style="bluefont"}
+absolute
+:::
+
+benefits for patients with higher baseline risk
+predictions, and followed the evolution of the stratified estimates closely
+(Figure \ref{fig:gusto}). The constant treatment effect model had somewhat lower
+AIC compared to the linear interaction model (AIC:
 `r gustoPerformance %>% filter(method == "constant treatment effect") %>% pull(aic) %>% round()` versus 
@@
 ```{r gusto, cache=TRUE, echo=FALSE, fig.cap="Individualized absolute benefit predictions based on baseline risk when using a constant treatment effect approach, a linear interaction approach and RCS smoothing using 3 knots.
-Risk stratified estimates of absolute benefit are presented within quartiles of baseline risk as reference.", fig.show="hold", out.width = '100%'}
+Risk stratified estimates of absolute benefit are presented within quartiles of baseline risk as reference.
+95\\% confidence bands were generated using 10,000 bootstrap resamples, where the prediction model was refitted in each run to capture the uncertainty in baseline risk predictions.", fig.show="hold", out.width = '100%'}
 grid::grid.raster(tiff::readTIFF(here("figures/gusto.tiff")))
@@
 effect was truly constant. Moreover, the assumption of a constant treatment
-effect may often be too strong [@Rothwell1995; @Kent2016].  For example, infants
+effect may often be too strong [@Rothwell1995; @Kent2016].  ~~For example, infants
 at lower risk of bronchopulmonary dysplasia benefit relatively more from vitamin
@@
 [@Sussman2015]. Hence, a linear interaction between baseline risk and the effect
-of treatment may be the most sensible approach with moderate sample sizes.
+of treatment may be the most sensible approach with moderate sample sizes.~~
 
@@
 approach often may be reasonable.
+
+:::{custom-style="bluefont"}
+Our results can also be interpreted in terms of bias-variance trade-off. The
+increasingly complex models considered allow for more degrees of freedom which,
+in turn increase the variance of our absolute benefit estimates. However, as was
+clear in our simulations, this increased complexity did not always result in
+substantial decrease in bias, especially with lower sample sizes and weaker
+treatment effects. Consequently, in most scenarios the simpler linear
+interaction model achieved the best variance-bias balance and outperformed the
+more complex RCS methods, even in the presence of non-linearity in the true
+underlying relationship between baseline risk and treatment effect. Conversely,
+the simpler constant treatment effect model was often heavily biased and, despite
+its lower variance, was outperformed by the other methods in the majority of the
+considered scenarios.
+:::
 
@@
 such approaches is their sensitivity to overfitting, which may exaggerate the
-magnitude of predicted benefits. In a wide range of simulation settings, a
+magnitude of predicted benefits.
+
+:::{custom-style="bluefont"}
+However, this can often be mitigated using methods such as cross-validation or
+regularization that penalizes the effect of treatment-covariate
+interactions. Therefore, in the presence of a limited set of true strong
+treatment-covariate interactions and adequate sample size one would expect
+treatment effect modeling methods to outperform risk modeling methods in a
+head-to-head comparison. [GOES WITH PREVIOUS]
+
+Nevertheless, in many cases treatment effect modifiers
+are not known and the available sample size does not allow the exploration of a
+large number of interaction effects. In these cases, risk modeling approaches
+like the ones presented here can provide individualized benefit predictions that
+improve on the "one-size-fits-all" overall RCT result. [GOES WITH NEXT]
+:::
+
+In a wide range of simulation settings, a
 simpler risk modeling approach was consistently better calibrated for benefit
@@
 of empirical settings still needs to be explored.
+
+:::{custom-style="bluefont"}
+Finally, in all our simulation scenarios we assumed all covariates to be
+statistically independent. This can be viewed as a limitation of our extensive
+simulation study, as this often is not the case in practice. However, our aim
+was to evaluate how sample size and the discriminative ability of the prediction
+model affect the predictive performance of increasingly complex regression-based
+methods for individualizing treatment effects. In this regard, that statistical
+independence of the simulated covariates allows for cleaner interpretation of
+the results. [I don't like that!!]
+:::
 

### 2022-09-04 code/CalibrationBase.R

@@ -99,10 +99,10 @@ res <- gridExtra::grid.arrange(
       plot.title = element_markdown(),
       axis.title = element_blank(),
       legend.direction = "horizontal",
-      legend.position = c(.354, .93),
+      legend.position = c(.354, .925),
       legend.text = element_text(size = 7),
       legend.title = element_text(size = 8),
-      axis.text.x = element_blank()
+      # axis.text.x = element_blank()
     ),
   plotList[[2]] +
     theme(
@@ -110,7 +110,7 @@ res <- gridExtra::grid.arrange(
       plot.title = element_markdown(),
       axis.title = element_blank(),
       legend.position = "none",
-      axis.text.x = element_blank()
+      # axis.text.x = element_blank()
     ),
   plotList[[3]] +
     theme(
@@ -138,6 +138,11 @@ res <- gridExtra::grid.arrange(
       )
     ),
     rot = 90
+  ),
+  bottom = grid::textGrob(
+    "Method",
+    just = "center",
+    gp = grid::gpar(fontsize = 10)
   )
 )
 

### 2022-09-04 code/CreateBootstrapData.R

@@ -64,11 +64,13 @@ bootstrapConfidenceInterval <- function(x, data) {
 
   riskLinearPredictor <- predict(
     prediction,
-    newdata = bootData %>% mutate(treatment = 0) %>% data.frame()
+    newdata = bootData %>%
+      dplyr::mutate(treatment = 0) %>%
+      data.frame()
   )
 
   bootData <- bootData %>%
-    mutate(
+    dplyr::mutate(
       riskLinearPredictor = riskLinearPredictor
     )
   constantModel <- fitModelBasedHte(
@@ -84,6 +86,23 @@ bootstrapConfidenceInterval <- function(x, data) {
     data     = bootData,
     settings = createRcsSettings()
   )
+
+  stratified <- fitStratifiedHte(
+      data     = bootData,
+      settings = createStratifiedSettings()
+  )
+
+  bootBenefitPredictionsStratified <- stratified$data %>%
+    dplyr::rename(
+      c(
+        "value" = "estimate",
+        "risk" = "meanRisk"
+      )
+    ) %>%
+    dplyr::mutate(
+      method = "stratified"
+    ) %>%
+    dplyr::select(riskStratum, value, risk, method)
 
   bootBenefitPredictionsConstant <- tibble(
     value = predictBenefitModelBasedHte(
@@ -114,7 +133,8 @@ bootstrapConfidenceInterval <- function(x, data) {
     bind_rows(
       bootBenefitPredictionsConstant,
       bootBenefitPredictionsLinear,
-      bootBenefitPredictionsRcs
+      bootBenefitPredictionsRcs,
+      bootBenefitPredictionsStratified
     ) %>%
       mutate(run = x)
   )
@@ -127,7 +147,7 @@ ParallelLogger::clusterRequire(cl, "rms")
 ParallelLogger::clusterRequire(cl, "SmoothHte")
 
 res <- ParallelLogger::clusterApply(
-  x = 1:1e3,
+  x = 1:1e4,
   cl = cl,
   fun = bootstrapConfidenceInterval,
   data = gusto
@@ -135,18 +155,38 @@ res <- ParallelLogger::clusterApply(
 
 mergedResults <- dplyr::bind_rows(res)
 
-cis <- mergedResults %>%
+mergedResultsNoStrat <- mergedResults %>%
+  filter(is.na(riskStratum)) %>%
+  select(-riskStratum)
+
[…]

### 2022-09-04 code/DiscriminationBase.R

@@ -85,19 +85,19 @@ res <- gridExtra::grid.arrange(
       plot.title = element_markdown(),
       axis.title = element_blank(),
       legend.direction = "horizontal",
-      legend.position = c(.364, .94),
+      legend.position = c(.364, .935),
       legend.text = element_text(size = 7),
       legend.title = element_text(size = 8),
-      panel.grid.minor = element_blank(),
-      axis.text.x = element_blank()
+      panel.grid.minor = element_blank()
+      # axis.text.x = element_blank()
     ),
   plotList[[2]] +
     theme(
       plot.title = element_markdown(),
       axis.title = element_blank(),
       legend.position = "none",
-      panel.grid.minor = element_blank(),
-      axis.text.x = element_blank()
+      panel.grid.minor = element_blank()
+      # axis.text.x = element_blank()
     ),
   plotList[[3]] +
     theme(
@@ -116,7 +116,12 @@ res <- gridExtra::grid.arrange(
   heights = c(1, 1.05),
   nrow = 2,
   ncol = 2,
-  left = "C-statistic for benefit"
+  left = "C-statistic for benefit",
+  bottom = grid::textGrob(
+    "Method",
+    just = "center",
+    gp = grid::gpar(fontsize = 10)
+  )
 )
 
 fileName <- paste0(

### 2022-09-04 code/GustoPlot.R

@@ -105,10 +105,14 @@ scale_custom <- list(
 
 plot <- ggplot() +
   geom_pointrange(
-    data = stratified$data,
+    data = bootstrapData %>%
+      filter(method == "stratified") %>%
+      bind_cols(
+        tibble(value = stratified$data$estimate)
+      ),
     aes(
-      x     = meanRisk,
-      y     = estimate,
+      x     = risk,
+      y     = value,
       ymin  = lower,
       ymax  = upper,
       color = "Stratified"
@@ -121,7 +125,7 @@ plot <- ggplot() +
       label = "Constant treatment effect"
     ),
     aes(
-      x     = x, 
+      x     = x,
       color = label,
       linetype = label
     ),

### 2022-09-04 code/PlotRmse.R

@@ -146,12 +146,13 @@ gridList <- list(
       plot.title = element_markdown(size = 9),
       axis.title.x = ggplot2::element_blank(),
       axis.title.y = ggplot2::element_blank(),
-      axis.text.x = element_blank(),
+      # axis.text.x = element_blank(),
+      axis.text.x = element_text(size = 8),
       axis.text.y = element_text(size = 8),
       legend.direction = "horizontal",
       legend.title = element_text(size = 7.5),
       legend.text = element_text(size = 7),
-      legend.position = c(.273, .87)
+      legend.position = c(.273, .848)
     ),
   absolutePlots$plot[[1]] +
     ggtitle("Simulated absolute benefit in treated patients") +
@@ -174,7 +175,8 @@ gridList <- list(
       panel.grid.minor = element_blank(),
       axis.title.x = element_blank(),
       axis.title.y = element_blank(),
-      axis.text.x = element_blank(),
+      # axis.text.x = element_blank(),
+      axis.text.x = element_text(size = 8),
       axis.text.y = element_text(size = 8),
       ggside.line = element_blank(),
       ggside.rect = element_blank(),
@@ -189,7 +191,8 @@ gridList <- list(
       panel.grid.minor = element_blank(),
       plot.title = element_markdown(size = 9),
       axis.title = element_blank(),
-      axis.text.x = element_blank(),
+      # axis.text.x = element_blank(),
+      axis.text.x = element_text(size = 8),
       axis.text.y = element_text(size = 8),
       ggside.line = element_blank(),
       ggside.rect = element_blank(),
@@ -219,7 +222,8 @@ gridList <- list(
       panel.grid.minor = element_blank(),
       axis.title.x = element_blank(),
       axis.title.y = element_blank(),
-      axis.text.x = element_blank(),
+      # axis.text.x = element_blank(),
+      axis.text.x = element_text(size = 8),
       axis.text.y = element_text(size = 8),
       ggside.line = element_blank(),
       ggside.rect = element_blank(),
@@ -234,7 +238,8 @@ gridList <- list(
       panel.grid.minor = element_blank(),
       plot.title = element_markdown(size = 9),
       axis.title = element_blank(),
-      axis.text.x = element_blank(),
+      # axis.text.x = element_blank(),
+      axis.text.x = element_text(size = 8),
       legend.position = "none"
     ),
   absolutePlots$plot[[3]] +
@@ -259,7 +264,8 @@ gridList <- list(
[…]
